---
title: "CEP-008: Derived variable verification"
author: "cchsflow review (PR #165)"
date: "2026-02-12"
format:
  html:
    toc: true
    code-fold: true
---

## Overview

This document verifies the refactored `calculate_pct_time()` and `categorize_pct_time()`
derived variable functions from PR #165. The functions were refactored from
legacy `if_else2()` to `dplyr::case_when()`, and `calculate_pct_time_A()` was removed
in favour of a single unified function.

**Key design change:** PUMF midpoints (4.5 and 15 years) were moved from
hardcoded R logic to the worksheet via the `SDCGRES_cont` intermediate variable.
Both PUMF and master paths now feed continuous years into the same function.

```{r}
#| label: setup
#| message: false
#| warning: false

devtools::load_all(here::here())
library(haven)
library(dplyr)
```

## calculate_pct_time: scalar tests

### Born in Canada

Respondents born in Canada (born_in_canada = 1) should always return 100,
regardless of age or years_in_canada.

```{r}
#| label: test-born-in-canada

stopifnot(calculate_pct_time(27, 1, 4.5) == 100)
stopifnot(calculate_pct_time(80, 1, 0) == 100)
stopifnot(calculate_pct_time(1, 1, NA) == 100)
cat("PASS: born in Canada always returns 100\n")
```

### Continuous years calculation

For immigrants (born_in_canada = 2), percent time = years_in_canada / age * 100.

```{r}
#| label: test-continuous-calculation

# PUMF midpoint: 0-9 years category (midpoint 4.5)
result <- calculate_pct_time(20, 2, 4.5)
stopifnot(result == 22.5)

# PUMF midpoint: 10+ years category (midpoint 15)
result <- calculate_pct_time(20, 2, 15)
stopifnot(result == 75)

# Master continuous: exact years
result <- calculate_pct_time(20, 2, 10)
stopifnot(result == 50)

result <- calculate_pct_time(40, 2, 30)
stopifnot(result == 75)

cat("PASS: continuous calculation correct for PUMF midpoints and master years\n")
```

### Missing data handling

Invalid or missing inputs should return `tagged_na("b")`.

```{r}
#| label: test-missing-data

# Age out of range
stopifnot(is_tagged_na(calculate_pct_time(-1, 2, 4.5), "b"))
stopifnot(is_tagged_na(calculate_pct_time(0, 2, 4.5), "b"))

# Invalid immigrant status
stopifnot(is_tagged_na(calculate_pct_time(20, 3, 4.5), "b"))
stopifnot(is_tagged_na(calculate_pct_time(20, 0, 4.5), "b"))

# NA inputs
stopifnot(is_tagged_na(calculate_pct_time(NA, 2, 4.5), "b"))
stopifnot(is_tagged_na(calculate_pct_time(20, NA, 4.5), "b"))
stopifnot(is_tagged_na(calculate_pct_time(20, 2, NA), "b"))
stopifnot(is_tagged_na(calculate_pct_time(NA, NA, NA), "b"))

cat("PASS: all invalid/missing inputs return tagged_na('b')\n")
```

### Output bounds validation

Values outside [0, 100] indicate inconsistent inputs (e.g., years_in_canada > age).
These are treated as data quality errors and return `tagged_na("b")`.

```{r}
#| label: test-output-bounds

# years_in_canada > age → impossible percentage (>100)
stopifnot(is_tagged_na(calculate_pct_time(20, 2, 25), "b"))

# Exact boundary: years_in_canada == age → 100% (valid)
stopifnot(calculate_pct_time(20, 2, 20) == 100)

# Just under boundary
result <- calculate_pct_time(20, 2, 19)
stopifnot(result == 95)

# Vector with out-of-bounds element
result <- calculate_pct_time(
  age = c(20, 30, 10),
  born_in_canada = c(2, 2, 2),
  years_in_canada = c(10, 35, 5)
)
stopifnot(result[1] == 50)
stopifnot(is_tagged_na(result[2], "b"))  # 35/30 = 116.7% → invalid
stopifnot(result[3] == 50)

cat("PASS: output bounds validation catches impossible percentages\n")
```

## calculate_pct_time: vectorised tests

The refactored function must work with vector inputs (not just row-by-row via
`rec_with_table()`). The original `if()`/`return()` pattern broke vectorisation.

```{r}
#| label: test-vectorised

result <- calculate_pct_time(
  age = c(20, 40, 30, 25, NA),
  born_in_canada = c(2, 1, 2, 2, 2),
  years_in_canada = c(4.5, 4.5, 15, 10, 5)
)

stopifnot(length(result) == 5)
stopifnot(result[1] == 22.5)           # immigrant, 4.5/20 * 100
stopifnot(result[2] == 100)            # born in Canada
stopifnot(result[3] == 50)             # immigrant, 15/30 * 100
stopifnot(result[4] == 40)             # immigrant, 10/25 * 100
stopifnot(is_tagged_na(result[5], "b")) # NA age

cat("PASS: vectorised input produces correct element-wise results\n")
```

### Mixed vector with all pathways

```{r}
#| label: test-mixed-vector

result <- calculate_pct_time(
  age = c(30, 50, -1, 20, 40),
  born_in_canada = c(1, 2, 2, 3, 2),
  years_in_canada = c(10, 25, 5, 10, NA)
)

stopifnot(result[1] == 100)              # born in Canada
stopifnot(result[2] == 50)               # 25/50 * 100
stopifnot(is_tagged_na(result[3], "b"))   # negative age
stopifnot(is_tagged_na(result[4], "b"))   # invalid born_in_canada = 3
stopifnot(is_tagged_na(result[5], "b"))   # NA years_in_canada

cat("PASS: mixed vector with all pathways\n")
```

## categorize_pct_time: categorical binning

The categorical function bins continuous percent time into deciles ("1" through
"10"), with "NA(a)" for not applicable and "NA(b)" for missing/invalid.

```{r}
#| label: test-cat-boundaries

# Boundary values for each bin
boundaries <- data.frame(
  input = c(0, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50,
            55, 60, 65, 70, 75, 80, 85, 90, 95, 100),
  expected = c("1", "1", "1", "2", "2", "3", "3", "4", "4", "5", "5",
               "6", "6", "7", "7", "8", "8", "9", "9", "10", "10")
)

results <- categorize_pct_time(boundaries$input)
mismatches <- boundaries[results != boundaries$expected, ]

if (nrow(mismatches) > 0) {
  cat("FAIL: mismatches found\n")
  print(mismatches)
  stop("Boundary test failed")
} else {
  cat("PASS: all", nrow(boundaries), "boundary values correctly binned\n")
}
```

### Special values

```{r}
#| label: test-cat-special

# tagged_na("a") -> "NA(a)"
stopifnot(categorize_pct_time(tagged_na("a")) == "NA(a)")

# tagged_na("b") -> "NA(b)"
stopifnot(categorize_pct_time(tagged_na("b")) == "NA(b)")

# Out of range -> "NA(b)"
stopifnot(categorize_pct_time(-1) == "NA(b)")
stopifnot(categorize_pct_time(101) == "NA(b)")

# Vector
result <- categorize_pct_time(c(5, 25, 55, 85, 100))
stopifnot(identical(result, c("1", "3", "6", "9", "10")))

cat("PASS: special values and vector input handled correctly\n")
```

## Return type verification

The refactored functions must return consistent types — `calculate_pct_time` returns
numeric (with `tagged_na` for missing), `categorize_pct_time` returns character.

```{r}
#| label: test-return-types

# calculate_pct_time returns numeric
r1 <- calculate_pct_time(20, 2, 10)
stopifnot(is.numeric(r1))

r2 <- calculate_pct_time(20, 1, 10)
stopifnot(is.numeric(r2))

r3 <- calculate_pct_time(NA, 2, 10)
stopifnot(is.numeric(r3))
stopifnot(is.na(r3))

# categorize_pct_time returns character
r4 <- categorize_pct_time(50)
stopifnot(is.character(r4))

r5 <- categorize_pct_time(-1)
stopifnot(is.character(r5))

r6 <- categorize_pct_time(tagged_na("a"))
stopifnot(is.character(r6))

cat("PASS: return types are consistent (numeric and character)\n")
```

## haven tagged_na verification

Confirm that `tagged_na("b")` is correctly produced and detectable. This
validates the haven integration — `tagged_na` values are NA to `is.na()` but
carry a tag accessible via `is_tagged_na()`.

```{r}
#| label: test-tagged-na

missing_result <- calculate_pct_time(NA, 2, 4.5)

# It is NA
stopifnot(is.na(missing_result))

# It is specifically tagged_na("b")
stopifnot(haven::is_tagged_na(missing_result, "b"))

# It is not tagged_na("a")
stopifnot(!haven::is_tagged_na(missing_result, "a"))

# Numeric comparison with tagged_na works
stopifnot(identical(missing_result, tagged_na("b")))

cat("PASS: haven tagged_na correctly produced and detectable\n")
```

## Summary

```{r}
#| label: summary

cat("\n=== ALL VERIFICATION TESTS PASSED ===\n\n")
cat("Functions verified:\n")
cat("  calculate_pct_time(age, born_in_canada, years_in_canada)\n")
cat("  categorize_pct_time(pct_time_der)\n\n")
cat("Test categories:\n")
cat("  - Born-in-Canada pathway (returns 100)\n")
cat("  - Continuous calculation (PUMF midpoints and master years)\n")
cat("  - Missing data handling (tagged_na('b'))\n")
cat("  - Output bounds validation (>100% → tagged_na('b'))\n")
cat("  - Vectorised operation (all pathways in single call)\n")
cat("  - Categorical binning (21 boundary values)\n")
cat("  - Special values (tagged_na('a'), out of range)\n")
cat("  - Return type consistency (numeric, character)\n")
cat("  - haven tagged_na integration\n")
```
