---
title: "CSV Validation and Cleaning"
author: "cchsflow Development Team"
date: "`r Sys.Date()`"
format: html
vignette: >
  %\VignetteIndexEntry{CSV Validation and Cleaning}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `standardise_csv()` function eliminates git merge conflicts in CSV files by fixing invisible Excel corruption issues. It's designed for Excel-first workflows with zero learning curve, plus optional enhanced validation for research collaboration.

## Quick Start

```{r setup, include=FALSE}
library(cchsflow)
```

### Basic Usage

The most common use case is fixing git conflicts in `variables.csv` or `variable_details.csv`:

```{r basic-usage, eval=FALSE}
# Fix git conflicts immediately
standardise_csv("inst/extdata/variables.csv")

# Check what would be fixed first (safe exploration)
result <- standardise_csv("inst/extdata/variables.csv", validate_only = TRUE)
if (!result$valid) {
  print(result$issues_remaining)
}
```

## Common Issues Fixed

The function automatically fixes:

- **UTF-8 BOM** (invisible bytes that break scripts)
- **Windows/Mac line endings** causing git conflicts
- **Inconsistent encoding** between team members
- **File formatting differences** from different editors
- **Column order inconsistencies** (reorders to schema-defined standard)
- **Missing/additional columns** (reports and preserves user extensions)

## Operating Modes

### Basic Mode (Default)

Perfect for Excel users who just want git conflicts to disappear:

```{r basic-mode, eval=FALSE}
# Basic mode - just fixes issues
result <- standardise_csv("variables.csv", collaboration = FALSE)
```

**What it does:**
- Removes UTF-8 BOM
- Fixes line endings (CRLF → LF, CR → LF)
- Standardizes column order
- Validates required fields are present

**What it doesn't do:**
- Strict schema validation
- Enum constraint checking
- Cross-field dependency validation

### Collaboration Mode

For research-quality validation before sharing with team:

```{r collaboration-mode, eval=FALSE}
# Collaboration mode - enhanced validation
result <- standardise_csv("variables.csv", collaboration = TRUE)
```

**Additional checks:**
- Schema-based validation
- Enum constraint validation (e.g., variableType must be "Categorical" or "Continuous")
- Cross-field dependencies (e.g., continuous variables should have units)
- Column completeness reporting

## Validation Examples

### Check Before Fixing

Always a good idea to see what would change:

```{r validate-only, eval=FALSE}
result <- standardise_csv("variables.csv", validate_only = TRUE)

if (result$valid) {
  cat("✅ File is clean!\n")
} else {
  cat("⚠️ Issues found:\n")
  for (issue in result$issues_remaining) {
    cat("  -", issue, "\n")
  }
}
```

### Understanding Return Values

```{r return-values, eval=FALSE}
result <- standardise_csv("variables.csv")

# Check what was done
str(result)
# List with components:
#   $success         - TRUE if standardisation succeeded
#   $mode            - "basic" or "collaboration"
#   $issues_fixed    - Character vector of problems resolved
#   $issues_remaining - Character vector of issues needing attention
#   $valid           - TRUE if file passes validation
```

## Working with Project-Specific Files

The function automatically recognizes project naming conventions:

```{r project-files, eval=FALSE}
# All these work automatically
standardise_csv("variables_MyProject.csv")
standardise_csv("ProjectName_variable_details.csv")
standardise_csv("variables_DemPoRT.csv")
standardise_csv("MyProject_variables.csv")
```

## Team Workflow Example

Here's a recommended workflow for teams collaborating on CSV files:

```{r team-workflow, eval=FALSE}
# Before editing in Excel
standardise_csv("variables.csv")

# ... edit in Excel ...

# After editing, before committing
result <- standardise_csv("variables.csv",
                          collaboration = TRUE,
                          validate_only = TRUE)

if (result$valid) {
  # Apply fixes
  standardise_csv("variables.csv", collaboration = TRUE)

  # Commit
  system("git add inst/extdata/variables.csv")
  system('git commit -m "Update variables.csv"')
} else {
  cat("⚠️ Please fix these issues before committing:\n")
  for (issue in result$issues_remaining) {
    cat("  -", issue, "\n")
  }
}
```

## Handling Specific Issues

### UTF-8 BOM (Excel Corruption)

Excel on Windows often adds an invisible UTF-8 BOM (Byte Order Mark) that breaks R scripts:

```{r bom-example, eval=FALSE}
# Before: File starts with invisible EF BB BF bytes
# After: Clean UTF-8 file

result <- standardise_csv("variables.csv")
# issues_fixed: "Removed UTF-8 BOM"
```

### Line Ending Issues

Different operating systems use different line endings:

- **Windows**: CRLF (`\r\n`)
- **Mac/Linux**: LF (`\n`)
- **Old Mac**: CR (`\r`)

```{r line-endings, eval=FALSE}
# Automatically converts all to LF
result <- standardise_csv("variables_from_windows.csv")
# issues_fixed: "Converted CRLF to LF line endings"
```

### Column Order

Keeps columns in schema-defined order:

```{r column-order, eval=FALSE}
# Before: variableType,variable,label,databaseStart
# After:  variable,label,variableType,databaseStart

result <- standardise_csv("variables.csv")
# issues_fixed: "Standardised column order"
```

### Missing Required Fields

```{r missing-fields, eval=FALSE}
# Validates that required fields are present
result <- standardise_csv("incomplete_variables.csv",
                          collaboration = TRUE,
                          validate_only = TRUE)

if (!result$valid) {
  # issues_remaining: "Missing required fields: variableType"
}
```

## Testing Your Changes

### Manual Testing

Create a test file with issues and verify they're fixed:

```{r manual-test, eval=FALSE}
# Create test file with CRLF line endings
test_content <- "variable,label,variableType\ntest1,Test,Categorical\n"
test_content <- gsub("\n", "\r\n", test_content)
test_file <- tempfile(fileext = ".csv")
writeLines(test_content, test_file)

# Fix it
result <- standardise_csv(test_file)

# Verify
file_content <- readChar(test_file, file.info(test_file)$size)
stopifnot(!grepl("\r", file_content))  # No CR characters remain
cat("✅ Line endings fixed successfully\n")
```

### Integration Testing with Real Files

Test on your actual CSV files:

```{r integration-test, eval=FALSE}
# Test on real files
files_to_test <- c(
  "inst/extdata/variables.csv",
  "inst/extdata/variable_details.csv"
)

for (file in files_to_test) {
  if (file.exists(file)) {
    cat("Testing", basename(file), "... ")
    result <- standardise_csv(file,
                              collaboration = TRUE,
                              validate_only = TRUE)
    if (result$valid) {
      cat("✅ PASS\n")
    } else {
      cat("⚠️ Issues found:\n")
      print(result$issues_remaining)
    }
  }
}
```

## Troubleshooting

### Schema Not Found

If you see "Schema file not found" errors:

```{r schema-troubleshooting, eval=FALSE}
# Check schema files exist
schema_files <- c(
  "inst/metadata/schemas/core/variables.yaml",
  "inst/metadata/schemas/core/variable_details.yaml"
)

for (schema in schema_files) {
  cat(schema, ":", ifelse(file.exists(schema), "✅", "❌"), "\n")
}

# If missing, collaboration mode won't work - use basic mode
result <- standardise_csv("variables.csv", collaboration = FALSE)
```

### File Modified But Still Has Issues

If the file was modified but still shows issues:

```{r still-has-issues, eval=FALSE}
# Run multiple times - some fixes may require multiple passes
result1 <- standardise_csv("variables.csv")
result2 <- standardise_csv("variables.csv")

# After second pass, check if any fixes were needed
if (length(result2$issues_fixed) == 0) {
  cat("✅ File is fully standardised\n")
}
```

### Git Still Showing Large Diffs

After standardization, git may show large diffs the first time:

```{r git-diffs, eval=FALSE}
# This is normal! The file is being fixed.
# After this commit, future edits will have clean, small diffs

# To see what changed:
system("git diff inst/extdata/variables.csv | head -20")

# Commit the standardised file
system("git add inst/extdata/variables.csv")
system('git commit -m "Standardise variables.csv formatting"')
```

## Advanced Usage

### Custom Output Path

Save to a different location:

```{r custom-output, eval=FALSE}
result <- standardise_csv(
  file_path = "messy_variables.csv",
  output_path = "clean_variables.csv"
)
```

### Batch Processing

Process multiple files:

```{r batch-process, eval=FALSE}
csv_files <- list.files("inst/extdata", pattern = "\\.csv$", full.names = TRUE)

results <- lapply(csv_files, function(file) {
  cat("Processing", basename(file), "... ")
  result <- standardise_csv(file, collaboration = TRUE)
  cat(ifelse(result$success, "✅\n", "⚠️\n"))
  result
})

# Summary
success_count <- sum(sapply(results, function(r) r$success))
cat(sprintf("\n%d/%d files processed successfully\n",
            success_count, length(results)))
```

### Pre-commit Hook Integration

Add to `.git/hooks/pre-commit`:

```bash
#!/bin/bash
# Automatically standardise CSV files before committing

Rscript -e '
csv_files <- c(
  "inst/extdata/variables.csv",
  "inst/extdata/variable_details.csv"
)

for (file in csv_files) {
  if (file.exists(file)) {
    result <- cchsflow::standardise_csv(file, collaboration = TRUE)
    if (!result$success) {
      cat("ERROR: CSV standardisation failed for", file, "\n")
      quit(status = 1)
    }
  }
}
'
```

## Best Practices

1. **Run before editing**: Always standardise before opening in Excel
2. **Run after editing**: Always standardise before committing
3. **Use validate_only first**: Check what will change before applying fixes
4. **Use collaboration mode for PR reviews**: Extra validation catches more issues
5. **Commit standardisation separately**: First commit fixes formatting, then commit content changes
6. **Team agreement**: Ensure all team members run standardisation

## See Also

- `?standardise_csv` - Function documentation
- `vignette("how_to_add_variables")` - Adding new variables
- `vignette("missing_value_conventions")` - Handling missing data
- CSV schemas: `inst/metadata/schemas/core/`

## Support

If you encounter issues:

1. Check this vignette's troubleshooting section
2. Run tests: `testthat::test_file("tests/testthat/test-csv-validate-and-clean.R")`
3. Report issues: https://github.com/Big-Life-Lab/cchsflow/issues
