---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1 - using cchsflow}     
  %\VignetteEngine{knitr::rmarkdown}  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

### Load packages

```{r eval= FALSE}
# Install release version from CRAN
install.packages("cchsflow")

# Install the most recent version from GitHub
devtools::install_github("Big-Life-Lab/cchsflow")
```

```{r results= 'hide', message = FALSE, warning=FALSE}
library(cchsflow)
```

The vignettes use the `bind_rows` function in the
[dplyr](https://www.rdocumentation.org/packages/dplyr/versions/0.7.8) package to
combine datasets.

```{r results = 'hide', message = FALSE, warning =FALSE}
library(dplyr)
```

# Transform variables into harmonized versions

Use `rec_with_table()` to transform the variables of a CCHS dataset.

## Example 1. Transform a single variable from a single database

In this example, the `sex` variable in the 2001 CCHS cycle is transformed into a
harmonized `sex` variable that is consistent all CCHS cycles.

```{r}
sex2001 <- rec_with_table(cchs2001_p, "DHH_SEX",
  log = TRUE,
  var_labels = c(DHH_SEX = "SEX")
)
```

```{r, echo=FALSE}
head(sex2001)
```

`cchs2001_p` is the name of test CCHS data that is included in the _cchsflow_
library. _cchsflow_ includes test CCHS data for the public use microdata file
(PUMF) for all general household survey (.1 versions) between 2001 and 2013.
Each survey cycle has a sample of 200 respondents. `cchs2000_p` indicates this 
isthe CCHS 2001 PUMF file. See [Reference](../reference/index.html) for more
information of each CCHS cycle.

You'll want to use _cchsflow_ on your own CCHS data that you'll need to first
load: something like `cchs2001 <- read.csv("~/data/cchs2001.csv")`. CCHS PUMF 
data can be obtained from Statistics Canada. 

`DHH_SEX` is the name of the harmonized variable for sex. Up to 2007, the
variable names in CCHS changed every cycle. Since 2007, there has _generally_
been a consistent name. _cchsflow_ uses the variable name from the 2007 cycle
whenever possible. For example, in the CCHS 2001, the variable for sex is
`DHHA_SEX` and in CCHS 2007 the same variable is `DHH_SEX`. This name, `DHH_SEX`
is the harmonized name for sex, meaning `DHHA_SEX` is transformed to `DHH_SEX`
using `rec_with_table()`.

The rules for the transformation and harmonization are in two dataframes that
are included in `variables` and `variable_details`, both included in _cchsflow_.
There are vignettes that further describe `variables` and `variable_details`,
including how to add or customize transformed variables.

## Which variables are transformed? How "good" is the transformation?

_cchsflow_ includes commonly used sociodemographic, health behaviour and health
status variables. In general, variables across CCHS cycles can be grouped into
five categories:

1. Consistent and unchanged across all cycles. These variables have no changes
   to the wording of the question that was asked, the response values and who
   was asked the question (same inclusion criteria and skip pattern).

2. Mostly consistent across cycles. These variables may have a small change in
   wording or response category. Notes for these variables are printed to the
   console when they are recoded with `rec_with_table()`. The notes are
   contained in `variable_details$notes`. Caution is required when using these
   variables. The notes should be reviewed to assess whether these variables can
   be used in your study. See [Example 2](#example2).

3. Either not collected in all cycles or have major changes between cycles.
   These variable may be included in _cchsflow_ with a note that says which
   cycles include the variable and can be transformed to a common variable. See
   [Example 4](#example4).

4. Mostly consistent but with complex changes or transformation rules. See the
   [Reference](../reference/index.html) for these derived and "Potentially
   problematic variables". See [Example 7](#example7).

5. Not consistent between cycles. Variable that are inconsistent or not asked in
   most cycles are not included in _cchsflow_. In the future, we hope to include
   these variables with an marker that they've been reviewed but rejected from
   inclusion in _cchsflow_.

## Example 2. Variable transfomation with a note {#example2}

`notes` provide context for the decisions that informed harmonization that may
affect your decision to use a harmonized variable.

```{r}
warning(
  "cchsflow includes ", length(unique(variable_details$notes)),
  " variables with notes."
)
```

For example, for `DHHGAGE_A` the category `NA::b` has the following `note`:

`Not applicable, don't know, refusal, not stated (96-99) were options only in CCHS 2003, but had zero responses`.

This note informs the decision to combine values 96 to 99 in `DHHGAGE_A` for
CCHS2003 into one common missing value `NA::b`. The label for the `NA::b`
category is, "`don't know (97); refusal (98); not stated (99)`".

By default, `rec_with_table()` prints notes to the console.

```{r}
BMI2001 <- rec_with_table(cchs2001_p, "HWTGBMI", log = TRUE)
```

### Notes with guidance or additional infomation

Several variables such as BMI have a quite a few notes reflecting a range of
changes that occurred over CCHS cycles. Variables with notes can usually be used
after reviewing and considering whether or how the variable changes affect your
use. There can also be further guidance in notes. For example, BMI notes include
advice to use `HWTGBMI_der`. Despite the large number of changes, BMI can be
used for most studies because the common transformed variable `HWTGBMI_der` was
derived from height and weight variables that are included and largely unchanged
across CCHS cycles. However, to use `HWTGBMI_der` you'll also need to include
the transformed measures for height (`HWTGHTM`) and weight (`HWTGWTK`). More
information about `HWTGBMI_der` can be found in the documentation reference for
'derived variable functions' or by typing `?BMI_fun` in the R console.

```{r}
BMI2001C <- rec_with_table(cchs2001_p,
  c("HWTGHTM", "HWTGWTK", "HWTGBMI_der"),
  log = TRUE
)
```

### Warning messages

Warning messages will appear when your dataset is missing variables in your
`variables` worksheet or `rec_with_table()` call.

### Variable names and labels included

`rec_with_table()` can include variable and value labels for the harmonized
variables using the `var_labels` attribute. By default, `rec_with_table()` will
include all labels if then all variables in `variables` are recoded. See
[Example 6](#example6).

```{r, warning=FALSE}
sex2001 <- rec_with_table(cchs2001_p, "DHH_SEX",
  log = TRUE,
  var_labels = c("DHH_SEX" = "Sex")
)
```

Labels can then be used in plots and tables.

```{r fig1, fig.height = 5, fig.width = 7, warning=FALSE}

barplot(
  table(as_label(sex2001$DHH_SEX)),
  col = cm.colors(5),
  ylim = c(0, 140),
  legend = TRUE,
  main = get_label(sex2001$DHH_SEX),
  ylab = "Frequency (n)"
)
```

### Missing data

If you've been watching, you may have noticed that _cchsflow_ recodes the values
in factor or categorical variables for 6 ("not applicable"), 7 ("don’t know"),
and 8 ("refusal"), 9 ("not stated") into `tagged_NA` values NA(a) ("not
applicable") and NA (b) ("missing"). There are two reasons. First, the typical
uses of CCHS consider the two values "not applicable" and "missing" for most
situations. Second, R and other open languages typically support only one NA.

Be carefull when use use `NA` in your study. See
[`missing data (tagged_na)`](../articles/tagged_na_usage.html)for more
information.

```{r}
library(haven)
x <- c(1:5, tagged_na("a"), tagged_na("b"), tagged_na("b"), NA)
x
# Do you want to calculate mean by excluding 'missing', 'not applicable', or both?
sum(!is_tagged_na(x, "a"))
sum(!is_tagged_na(x, "b"))
sum(!is.na(x))
sum(!is_tagged_na(x))
```

### Default arguments for `rec_with_table()`

With `rec_with_table()` there are also default arguments that do not need to be
called unless you would like to modify them.

- `variables` defaults to NULL. This results in the `variables.csv` sheet in the
  cchsflow package being used to specify which variables to transform. This
  argument can be modified if you want use your own `variables.csv` sheet or by
  specifying particular variables to transform.
- `database_name` defaults to NULL. This results in `rec_with_table()` using the
  database indicated in the `data` argument.
- `variable_details` defaults to NULL. This results in the
  `variable_details.csv` sheet in the cchsflow being used. This argument can be
  modified if want use your own `variable_details.csv` sheet.
- `else_value` defaults to NA. Values out of range set to NA.
- `append_to_data` defaults to FALSE. Transformed variables will not be appended
  to the original CCHS dataset.
- `log` defaults to FALSE. Logs of the variable transformations will not be
  displayed.
- `notes` defaults to TRUE. Information in the `Notes` column from
  `variable_details.csv` will be displayed
- `var_labels` is set to NULL. This argument can be used to add labels to a
  transformed dataset that only contains a subset of variables from
  `variables.csv`. The format to add labels to a subset of variables is
  `c(variable = "Label")`.
- `attach_data_name` is set to FALSE. This argument can be used to append a
  column that denotes the name of the dataset each transformed row came from.
  See [Example 8](#example8).

## Example 3. Transform a single variable from multiple CCHS datasets

This example shows how you can transform and combine a variable across multiple
CCHS cycles. The sex variable in CCHS 2001 (`DHHA_SEX`) and CCHS 2012
(`DHH_SEX`) is transformed a common variable (`DHH_SEX`) and combined into a
single dataset. In _cchsflow_ the CCHS 2007 variable name is used as the common
variable name. See the [variable_details vignette](variable_details.html) for
more information.

```{r, warning = FALSE}
sex2001 <- rec_with_table(cchs2001_p, "DHH_SEX", log = TRUE)
head(sex2001)

sex2012 <- rec_with_table(cchs2012_p, "DHH_SEX", log = TRUE)
tail(sex2012)
```

```{r, warning=FALSE}
combined_sex <- bind_rows(sex2001, sex2012)
```

```{r, echo=FALSE}
head(combined_sex)
tail(combined_sex)
```

## Example 4. Transform a single variable from multiple databases that changes categories between cycles. {#example4}

There are many variables in the CCHS that changes in categories between cycles
in a way that the variable cannot preserve all categories for all CCHS cycles.
An example is the categorical variable for `age`. _cchsflow_ offers three
options to facilitate the use of these variables with inconsistent categories
across cycles.

### Option 1: transform category `age` variable into a common variable for only cycles with the same category responses

Transform the `age` variable into variables that cannot be combined across
cycles. The categories in `age` variable in the CCHS changed in 2005 and
therefore it is not possible to have the same `age` categories across all CCHS
cycles.

`DHHGAGE_A` is the age variable for CCHS cycles 2001-2003, and `DHHGAGE_B` is
the age variable for CCHS cycles 2005-2014.

```{r, warning=FALSE}
age2001 <- rec_with_table(cchs2001_p, "DHHGAGE_A")
head(age2001)

age2012 <- rec_with_table(cchs2012_p, "DHHGAGE_B")
tail(age2012)
```

```{r, warning=FALSE}
combined_age_cat <- bind_rows(age2001, age2012)
```

```{r, echo=FALSE}
head(combined_age_cat)
tail(combined_age_cat)
```

### Option 2: transform the categorical `age` variable into a continuous `age_cont` variable

Transform categorical variables such as `age` into a single harmonized
`age_cont` variable. This variable takes the midpoint age of each category for
all CCHS cycles. With this option, the age category variable from all CCHS
cycles can be combined into a single dataset.

```{r, warning=FALSE}
age2001_cont <- rec_with_table(cchs2001_p, "DHHGAGE_cont")
head(age2001_cont)

age2012_cont <- rec_with_table(cchs2012_p, "DHHGAGE_cont")
tail(age2012_cont)
```

```{r, warning= FALSE}
combined_age_cont <- bind_rows(age2001_cont, age2012_cont)
```

```{r, echo=FALSE}
head(combined_age_cont)
tail(combined_age_cont)
```

### Option 3: transform the categorical `age` variable into a harmonized categorical variable

_cchsflow_ often includes a common harmonized categorical variable that has new,
consistent categories across all cycles. These new categorical variables
generally have fewer category levels than the original variables, reflecting the
need to combine categories. Read the `Notes` variable for details.

### Label variables and categories

Use `set_data_labels()` to label the variables in your final dataset. Labels are
lost during the database merging. `set_data_labels()` set labels with the
orginal information in `variables` and `variable_details`.

```{r}
labeled_combined_cchs <- set_data_labels(
  data_to_label = combined_age_cont,
  variable_details = variable_details,
  variables_sheet = variables
)
```

## Example 5. Transform multiple variables from multiple datasets

The variables argument in `rec_with_table()` allows multiple variables to be
transformed from a CCHS dataset. In this example, the age and sex variables from
the 2001 and 2012 CCHS datasets will be transformed and labeled using
`rec_with_table()`. They will then be combined into a single dataset and labeled
using `set_data_labels()`.

```{r, warning=FALSE}
age_sex_2001 <- rec_with_table(cchs2001_p, c("DHHGAGE_cont", "DHH_SEX"),
  var_labels = c(DHHGAGE_cont = "Age", DHH_SEX = "Sex")
)

get_label(age_sex_2001)

age_sex_2012 <- rec_with_table(cchs2012_p, c("DHHGAGE_cont", "DHH_SEX"),
  var_labels = c(DHHGAGE_cont = "Age", DHH_SEX = "Sex")
)

get_label(age_sex_2012)
```

In the above example, `var_labels` is called in `rec_with_table()` to label the
age and sex variables in the 2001 and 2012 datasets. Use `get_label()` to view
the variable labels in your transformed dataset. As mentioned previously,
`var_labels` can be used all the variables in `variables.csv` or a subset of
variables.

```{r, warning=FALSE}
combined_age_sex <- bind_rows(age_sex_2001, age_sex_2012)
labeled_combined_age_sex <- set_data_labels(
  data_to_label = combined_age_sex,
  variable_details = variable_details, variables_sheet = variables
)
```

In the above code, `set_data_labels()` is used to label the combined age and sex
dataset. Similar to before, you can check if labels have been added using
`get_label()`.

```{r, warning=FALSE}
get_label(labeled_combined_age_sex)
```

For more information on `get_label()` and other label helper functions, please
refer to the [sjlabelled](https://strengejacke.github.io/sjlabelled/reference/)
package.

## Example 6. Transform all variables in the variable_details sheet {#example6}

All the variables listed in `variables.csv` & `variable_details.csv` will be
transformed if the variables argument in `rec_with_table()` is not specified. In
this example, all variables specified in the two worksheets will be transformed,
combined, and labeled for the 2001 and 2012 CCHS cycles.

```{r, echo=FALSE}
options(htmlwidgets.TOJSON_ARGS = list(na = "string"))
```

```{r, warning=FALSE}
transformed2001 <- rec_with_table(data = cchs2001_p, notes = FALSE)

transformed2012 <- rec_with_table(data = cchs2012_p, notes = FALSE)

combined_cchs <- bind_rows(transformed2001, transformed2012)
labeled_combined_cchs <- set_data_labels(
  data_to_label = combined_cchs,
  variable_details = variable_details, variables_sheet = variables
)
```

## Example 7. Transform CCHS derived variables such as Body Mass Index (BMI) {#example7}

CCHS derived variables are recoded (harmonized) using the same method as other
CCHS _cchsflow_ variables, with one additional considerations.

Derived variables in CCHS are created from existing, original variables that are
transformed into a new variable. An example is body mass index (BMI).
Respondents are asked to report their height and weight. BMI is then derived
(calculated) using recoded variables for height and weight.

Recoding `BMI` into a harmonized variable requires their underlying initial
variables (height and weight). This first step of harmonizing underlying
variables is completed for you, but you must have the underlying variables in
your dataset.

Using the BMI example illustrated in the
[derived variables](derived_variables.html) article, `rec_with_table()` is able
to transform BMI across multiple cycles provided that height (`HWTGHTM`) and
weight (`HWTGWTK`) are specified.

```{r, warning=FALSE}
bmi2001 <- rec_with_table(cchs2001_p, c("HWTGHTM", "HWTGWTK", "HWTGBMI_der"),
  log = TRUE
)

head(bmi2001)

bmi2012 <- rec_with_table(cchs2012_p, c("HWTGHTM", "HWTGWTK", "HWTGBMI_der"),
  log = TRUE
)
tail(bmi2012)

combined_bmi <- bind_rows(bmi2001, bmi2012)

head(combined_bmi)
tail(combined_bmi)
```

## Example 8: Distinguishing which cycle data is from in combined datasets

It can be important to know from which cycle data comes from after combining all
the cycles. `rec_with_table()` provides the option of appending a column that
indicates from which dataset respondents come from by setting the parameter
`attach_data_name` to be TRUE. Using the BMI example in [Example 7](#example7),
`rec_with_table()` can specify which dataset each respondent came from.

```{r, warning=FALSE}
bmi2001 <- rec_with_table(cchs2001_p, c("HWTGHTM", "HWTGWTK", "HWTGBMI_der"),
  log = TRUE, attach_data_name = TRUE
)

head(bmi2001)

bmi2012 <- rec_with_table(cchs2012_p, c("HWTGHTM", "HWTGWTK", "HWTGBMI_der"),
  log = TRUE, attach_data_name = TRUE
)
tail(bmi2012)

combined_bmi <- bind_rows(bmi2001, bmi2012)

head(combined_bmi)
tail(combined_bmi)
```