---
title: "missing data (tagged_na)"
output: html_document
vignette: >
  %\VignetteIndexEntry{3 - variable_details.csv}     
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Missing data (`NA` --> `tagged_na`)

Base R supports only one type of NA ('not available') to represent missing 
values. The CCHS, however, has several types of missing data or incomplete 
category responses. `cchsflow` uses the `haven` package [`tagged_na()`](https://haven.tidyverse.org/reference/tagged_na.html) 
to allow multiple missing data categories. `tagged_na()` adds an addition 
character to the NA value, thereby allowing users to define additional missing 
data types. `tagged_na()` applies only for numeric values, as character base 
values can use any string to represent NA or missing data. 

## CCHS approach to coding missing data

CCHS category values can change across variables and survey 
cycles, but the most common values are:

| CCHS category value | Label |
|----------------|--------------|
| 6                  | not applicable   |
| 7                  | don't know       |
| 8                  | refusal          |
| 9                  | not stated       |


`cchsflow` recodes the four CCHS category missing data categories values into 
two NA values that are commonly used for most studies. NA(a) = 'not applicable' 
and NA(b) = 'missing'. Furthermore, variables may be entirely missing from a 
specific cycle, which results in 'not asked' missing data that is not included 
or coded in CCHS surveys, recoded to NA(a)

**Summary of `tagged_na` values and their corresponding CCHS category values.**

| cchsflow `tagged_na` | CCHS category value | Label |
|----------------|--------------|
| NA(a) | 6                  | not applicable     |
| NA(b) | 7                  | don't know         |
| NA(b) | 8                  | refusal            |
| NA(b) | 9                  | not stated         |
| NA(a) | question not asked in the survey cycle  |

## Example `haven::tagged_na()`
```{r cars}
library(haven)
x <- c(1:5, tagged_na("a"), tagged_na("z"), NA)
na_tag(x) # is used to read the tagged NA in most other functions they are still viewed as NA
print_tagged_na(x) # Is used to print the na as well as their tag
```

