---
title: "tagged_na_usage"
output: html_document
vignette: >
  %\VignetteIndexEntry{3 - variable_details.csv}     
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reasons for tagged_na usage

It is often useful to store different types of NA to represent wide variety of missing values.
Base R unfortunately does not support more then one type of NA. Different people have decided to handle this in their own manner.

For `cchsflow`, we have decided to go with the `haven` package for handling of missing. The haven package has an [introduction](https://haven.tidyverse.org/reference/tagged_na.html) of tagged_na. This is only applied to numeric values as character base values can use any string to represent NA. Haven allows you to attach an additional byte to the numeric NA value that can be distinguished. This also allows normal treatment of NA, as well as special distinction for missing values.

## Example
```{r cars}
library(haven)
x <- c(1:5, tagged_na("a"), tagged_na("z"), NA)
na_tag(x) # is used to read the tagged NA in most other functions they are still viewed as NA
print_tagged_na(x) # Is used to print the na as well as their tag
```

