---
title: "Checking and Fixing Worksheet Formatting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking and Fixing Worksheet Formatting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The cchsflow package maintains two critical CSV worksheets that define
variable metadata:

- `inst/extdata/variables.csv` - Variable definitions and metadata
- `inst/extdata/variable_details.csv` - Recoding rules for the variables

These worksheets are frequently modified by contributors using various tools
(Microsoft Excel, LibreOffice Calc, Notepad, etc.) across different operating
systems. This diversity can lead to formatting inconsistencies that make code
review difficult and error-prone.

This article explains how to use the `check_worksheet()` and `fix_worksheet()`
functions to detect and correct formatting issues in these worksheets.

## The Problem

When contributors edit the CSV worksheets with different tools, several
formatting issues can arise:

1. **Inconsistent Quoting**: Different CSV editors apply different quoting
   strategies. Some tools quote all fields regardless of content, while others
   only quote fields containing special characters (commas, quotes, newlines).

2. **Line Ending Variations**: Operating systems use different line ending
   conventions:
   - Unix/Linux/macOS use LF (`\n`)
   - Windows uses CRLF (`\r\n`)

3. **Structural Inconsistencies**: Manual editing can introduce structural
   problems:
   - Unsorted rows in the variables sheet
   - Empty header columns added accidentally
   - Column order being changed

4. **Poor Git Diffs**: These formatting differences cause Git to report that
   entire files have changed when semantically only a single cell was modified.
   This makes code review difficult.

## Formatting Standards

The worksheets are expected to follow these standards:

| Standard | Requirement |
|----------|-------------|
| Line Endings | LF (`\n`) only |
| Quoting | Minimal - only when required by CSV spec |
| Row Sorting | Rows sorted by `variable` column (variables sheet only) |
| Structure | No trailing empty columns |
| Column Order | Columns must follow the defined order |

## Checking for Errors

Use the `check_worksheet()` function to validate a worksheet against the
formatting standards:

```{r, eval=FALSE}
library(cchsflow)

# Check the variables sheet
errors <- check_worksheet(
  file_path = "inst/extdata/variables.csv",
  file_type = "variables"
)

# Check the variable details sheet
errors <- check_worksheet(
  file_path = "inst/extdata/variable_details.csv",
  file_type = "variable_details"
)
```

### Understanding the Output

The function returns a list of error objects. Each error object contains:

- `error_type`: The type of error detected
- `file_type`: Which worksheet type (`variables` or `variable_details`)
- `file_path`: Path to the file with the error
- `message`: A human-readable description of the error

Additional fields may be present depending on the error type (e.g., `row_num`,
`col_num`).

### Error Types

The checker detects the following error types:

| Error Type | Description |
|------------|-------------|
| `file_not_found` | The specified file does not exist |
| `invalid_csv` | The file is not valid CSV |
| `line_ending_crlf` | A row uses CRLF line endings instead of LF |
| `excessive_quoting` | A cell is quoted when it doesn't need to be |
| `column_order` | Columns are not in the expected order |
| `unsorted_rows` | Rows are not sorted by the `variable` column |
| `empty_columns` | Trailing empty columns exist |

## Fixing Errors

Use the `fix_worksheet()` function to automatically correct formatting issues:

```{r, eval=FALSE}
library(cchsflow)

# Fix the variables sheet
result <- fix_worksheet(
  file_path = "inst/extdata/variables.csv",
  file_type = "variables"
)

# Fix the variable details sheet
result <- fix_worksheet(
  file_path = "inst/extdata/variable_details.csv",
  file_type = "variable_details"
)
```

### Understanding the Output

The function returns a list of error objects, each with an added `fixed` field
indicating whether the error was corrected:

```{r, eval=FALSE}
result <- fix_worksheet("inst/extdata/variables.csv", "variables")

# Count fixed errors
fixed_count <- sum(sapply(result, function(e) e$fixed))
message(sprintf("Fixed %d errors", fixed_count))
```

### Which Errors Can Be Fixed?

| Error Type | Can Be Fixed? |
|------------|---------------|
| `file_not_found` | No |
| `invalid_csv` | No |
| `line_ending_crlf` | Yes |
| `excessive_quoting` | Yes |
| `column_order` | Yes |
| `unsorted_rows` | Yes (variables sheet only) |
| `empty_columns` | Yes |

## Command-Line Scripts

For convenience, the package includes command-line scripts that can be run
directly:

### Checking Worksheets

```bash
Rscript exec/check-worksheets.R
```

This script checks both worksheets and:

- Prints all errors found to the console
- Exits with a non-zero status code if any errors are found (useful for CI/CD)
- Exits with status code 0 if no errors are found

### Fixing Worksheets

```bash
Rscript exec/fix-worksheets.R
```

This script fixes both worksheets and prints a summary of all corrections
applied.

## CI/CD Integration

The checker script is designed to integrate with CI/CD pipelines. The
repository includes a GitHub Actions workflow that automatically checks the
worksheets on pull requests.

When a pull request modifies either CSV file, the workflow:

1. Runs the checker script
2. Fails the build if any formatting violations are detected
3. Reports the specific errors in the build log

This ensures that formatting issues are caught before they are merged into the
main branch.

## Workflow Recommendations

### Before Committing Changes

After editing either worksheet, run the fixer script to ensure consistent
formatting:

```bash
Rscript exec/fix-worksheets.R
git add inst/extdata/variables.csv inst/extdata/variable_details.csv
git commit -m "Update worksheet with fixed formatting"
```

### Reviewing Pull Requests

When reviewing pull requests that modify the worksheets:

1. Check that the CI workflow passed
2. Review only the semantic changes (the formatting should be consistent)
3. If the CI fails, ask the contributor to run the fixer script

## Summary

The `check_worksheet()` and `fix_worksheet()` functions help maintain
consistent formatting in the cchsflow worksheets. Using these tools ensures:

- Clean, semantic Git diffs
- Easier code review
- Consistent file formatting regardless of the editing tool used
- Automated enforcement through CI/CD integration
