---
title: "Sleep variable availability matrix"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  eval: true
---

## Overview

This document provides availability matrices for sleep variables across CCHS PUMF cycles, with specific focus on Ontario respondents by age group for dementia research.

**Key finding:** Sleep variables are **NOT available** in 2003, 2005, and 2009-2010 PUMF files.

## Setup

```{r setup}
library(haven)
library(dplyr)
library(tidyr)
library(knitr)
library(here)

# PUMF data location
pumf_dir <- "../../working_data_and_documentation/pumf-rdata"

# Helper function to load RData files
load_pumf_data <- function(file_path) {
  if (!file.exists(file_path)) return(NULL)
  env <- new.env()
  load(file_path, envir = env)
  get(ls(env)[1], envir = env)
}

# Sleep variable configuration by cycle
# NOTE: Province and age variable names differ by cycle!
cycles <- list(
  "2001" = list(
    file = "CCHS_2001.RData",
    hours = "GENA_03",
    trouble = "GENA_04",
    refreshing = "GENA_05",
    awake = "GENA_06",
    scale = "3-point",
    age_var = "DHHAGAGE",
    prov_var = "GEOAGPRV",
    ontario_code = 35,
    max_valid = 3
  ),
  "2007_2008" = list(
    file = "CCHS_2007_2008.RData",
    hours = "SLP_01",
    trouble = "SLP_02",
    refreshing = "SLP_03",
    awake = "SLP_04",
    scale = "5-point",
    age_var = "DHHGAGE",
    prov_var = "GEOGPRV",
    ontario_code = 35,
    max_valid = 5
  ),
  "2011_2012" = list(
    file = "CCHS_2011_2012.RData",
    hours = "SLPG01",
    trouble = "SLP_02",
    refreshing = "SLP_03",
    awake = "SLP_04",
    scale = "5-point",
    age_var = "DHHGAGE",
    prov_var = "GEOGPRV",
    ontario_code = 35,
    max_valid = 5
  ),
  "2013_2014" = list(
    file = "CCHS_2013_2014.RData",
    hours = "SLPG01",
    trouble = "SLP_02",
    refreshing = "SLP_03",
    awake = "SLP_04",
    scale = "5-point",
    age_var = "DHHGAGE",
    prov_var = "GEOGPRV",
    ontario_code = 35,
    max_valid = 5
  ),
  "2015_2016" = list(
    file = "CCHS_2015_2016.RData",
    hours = "SLPG005",
    trouble = "SLP_010",
    refreshing = "SLP_015",
    awake = "SLP_020",
    scale = "5-point",
    age_var = "DHHGAGE",
    prov_var = "GEO_PRV",
    ontario_code = 35,
    max_valid = 5
  ),
  "2017_2018" = list(
    file = "CCHS_2017_2018.RData",
    hours = "SLPG005",
    trouble = "SLP_010",
    refreshing = "SLP_015",
    awake = "SLP_020",
    scale = "5-point",
    age_var = "DHHGAGE",
    prov_var = "GEO_PRV",
    ontario_code = 35,
    max_valid = 5
  )
)

# Age cutoffs for dementia research
age_cutoffs <- c(50, 55, 60)

# Cycles confirmed to NOT have sleep variables
missing_cycles <- c("2003", "2005", "2009_2010")
```

## Availability matrix: Canada-wide

```{r canada_matrix}
generate_canada_matrix <- function(cycles, pumf_dir) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) {
      results[[cycle_name]] <- data.frame(
        cycle = cycle_name,
        scale = cfg$scale,
        total_n = NA,
        hours_n = NA,
        hours_pct = NA,
        trouble_n = NA,
        trouble_pct = NA
      )
      next
    }

    data <- load_pumf_data(file_path)
    total_n <- nrow(data)

    # Hours sleep (continuous variable, check for valid range 0-24)
    hours_n <- if (cfg$hours %in% names(data)) {
      sum(!is.na(data[[cfg$hours]]) & data[[cfg$hours]] >= 0 & data[[cfg$hours]] <= 24, na.rm = TRUE)
    } else 0

    # Trouble sleeping (categorical, valid codes 1-3 for 2001, 1-5 for 2007+)
    max_valid <- if (cfg$scale == "3-point") 3 else 5
    trouble_n <- if (cfg$trouble %in% names(data)) {
      sum(!is.na(data[[cfg$trouble]]) & data[[cfg$trouble]] >= 1 & data[[cfg$trouble]] <= max_valid, na.rm = TRUE)
    } else 0

    results[[cycle_name]] <- data.frame(
      cycle = cycle_name,
      scale = cfg$scale,
      total_n = total_n,
      hours_n = hours_n,
      hours_pct = round(100 * hours_n / total_n, 1),
      trouble_n = trouble_n,
      trouble_pct = round(100 * trouble_n / total_n, 1)
    )
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

canada_matrix <- generate_canada_matrix(cycles, pumf_dir)
kable(canada_matrix,
      caption = "Canada-wide sleep variable availability",
      col.names = c("Cycle", "Scale", "Total N", "Hours N", "Hours %", "Trouble N", "Trouble %"))
```

## Missing cycles confirmation

The following cycles do **NOT** have sleep variables in their PUMF files:

```{r missing_cycles_table}
missing_df <- data.frame(
  Cycle = c("2003", "2005", "2009-2010"),
  Status = rep("No sleep variables", 3),
  Verification = c(
    "DDI review: GENA_03-06 not present",
    "DDI review: SLP_* variables not present",
    "DDI review: SLP_* variables not present"
  )
)

kable(missing_df, caption = "Cycles without sleep variables (verified via DDI)")
```

## Availability matrix: Ontario by age group

```{r ontario_matrix}
generate_ontario_matrix <- function(cycles, pumf_dir, age_cutoffs) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) next

    data <- load_pumf_data(file_path)

    # Check if province and age variables exist
    if (!cfg$prov_var %in% names(data)) next
    if (!cfg$age_var %in% names(data)) next

    # Filter to Ontario
    ontario_data <- data[data[[cfg$prov_var]] == cfg$ontario_code, ]

    for (age_cut in age_cutoffs) {
      # Filter by age (DHHGAGE is grouped: 1=12-14, 2=15-17, ..., 16=80+)
      # Age 50+ corresponds to DHHGAGE >= 11 (50-54)
      # Age 55+ corresponds to DHHGAGE >= 12 (55-59)
      # Age 60+ corresponds to DHHGAGE >= 13 (60-64)
      age_code <- switch(as.character(age_cut),
        "50" = 11,
        "55" = 12,
        "60" = 13
      )

      age_filtered <- ontario_data[ontario_data[[cfg$age_var]] >= age_code, ]
      ontario_n <- nrow(age_filtered)

      # Hours sleep valid
      hours_valid <- if (cfg$hours %in% names(age_filtered) && nrow(age_filtered) > 0) {
        sum(!is.na(age_filtered[[cfg$hours]]) &
            age_filtered[[cfg$hours]] >= 0 &
            age_filtered[[cfg$hours]] <= 24, na.rm = TRUE)
      } else 0

      # Trouble sleeping valid
      max_valid <- if (cfg$scale == "3-point") 3 else 5
      trouble_valid <- if (cfg$trouble %in% names(age_filtered) && nrow(age_filtered) > 0) {
        sum(!is.na(age_filtered[[cfg$trouble]]) &
            age_filtered[[cfg$trouble]] >= 1 &
            age_filtered[[cfg$trouble]] <= max_valid, na.rm = TRUE)
      } else 0

      results[[paste0(cycle_name, "_", age_cut)]] <- data.frame(
        cycle = cycle_name,
        scale = cfg$scale,
        age_cutoff = paste0(age_cut, "+"),
        ontario_n = ontario_n,
        hours_valid = hours_valid,
        hours_pct = ifelse(ontario_n > 0, round(100 * hours_valid / ontario_n, 1), NA),
        trouble_valid = trouble_valid,
        trouble_pct = ifelse(ontario_n > 0, round(100 * trouble_valid / ontario_n, 1), NA)
      )
    }
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

ontario_matrix <- generate_ontario_matrix(cycles, pumf_dir, age_cutoffs)
kable(ontario_matrix,
      caption = "Ontario sleep variable availability by age group",
      col.names = c("Cycle", "Scale", "Age", "Ontario N", "Hours valid", "Hours %", "Trouble valid", "Trouble %"))
```

## Ontario summary by cycle

```{r ontario_summary}
if (exists("ontario_matrix") && nrow(ontario_matrix) > 0) {
  # Summarize by cycle for age 50+
  ontario_50plus <- ontario_matrix %>%
    filter(age_cutoff == "50+") %>%
    select(cycle, scale, ontario_n, hours_valid, trouble_valid)

  kable(ontario_50plus,
        caption = "Ontario 50+ respondents with valid sleep data",
        col.names = c("Cycle", "Scale", "Ontario 50+ N", "Hours valid", "Trouble valid"))
}
```

## Hours sleep distribution

```{r hours_distribution}
get_hours_summary <- function(cycles, pumf_dir) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) next

    data <- load_pumf_data(file_path)

    if (!cfg$hours %in% names(data)) next

    hours <- data[[cfg$hours]]
    hours_valid <- hours[!is.na(hours) & hours >= 1 & hours <= 12]

    results[[cycle_name]] <- data.frame(
      cycle = cycle_name,
      variable = cfg$hours,
      n_valid = length(hours_valid),
      mean_hours = round(mean(hours_valid), 1),
      median_hours = median(hours_valid),
      pct_lt6 = round(100 * sum(hours_valid < 6) / length(hours_valid), 1),
      pct_6to8 = round(100 * sum(hours_valid >= 6 & hours_valid <= 8) / length(hours_valid), 1),
      pct_gt8 = round(100 * sum(hours_valid > 8) / length(hours_valid), 1)
    )
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

hours_summary <- get_hours_summary(cycles, pumf_dir)
kable(hours_summary,
      caption = "Hours of sleep distribution by cycle",
      col.names = c("Cycle", "Variable", "Valid N", "Mean", "Median", "< 6h %", "6-8h %", "> 8h %"))
```

## Trouble sleeping distribution

```{r trouble_distribution}
get_trouble_summary <- function(cycles, pumf_dir) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) next

    data <- load_pumf_data(file_path)

    if (!cfg$trouble %in% names(data)) next

    trouble <- data[[cfg$trouble]]

    if (cfg$scale == "3-point") {
      # 2001: 1=Most of the time, 2=Sometimes, 3=Never
      valid <- trouble[!is.na(trouble) & trouble >= 1 & trouble <= 3]
      results[[cycle_name]] <- data.frame(
        cycle = cycle_name,
        variable = cfg$trouble,
        scale = cfg$scale,
        n_valid = length(valid),
        cat1_pct = round(100 * sum(valid == 1) / length(valid), 1),
        cat2_pct = round(100 * sum(valid == 2) / length(valid), 1),
        cat3_pct = round(100 * sum(valid == 3) / length(valid), 1),
        cat4_pct = NA,
        cat5_pct = NA
      )
    } else {
      # 2007+: 1=None, 2=A little, 3=Some, 4=Most, 5=All
      valid <- trouble[!is.na(trouble) & trouble >= 1 & trouble <= 5]
      results[[cycle_name]] <- data.frame(
        cycle = cycle_name,
        variable = cfg$trouble,
        scale = cfg$scale,
        n_valid = length(valid),
        cat1_pct = round(100 * sum(valid == 1) / length(valid), 1),
        cat2_pct = round(100 * sum(valid == 2) / length(valid), 1),
        cat3_pct = round(100 * sum(valid == 3) / length(valid), 1),
        cat4_pct = round(100 * sum(valid == 4) / length(valid), 1),
        cat5_pct = round(100 * sum(valid == 5) / length(valid), 1)
      )
    }
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

trouble_summary <- get_trouble_summary(cycles, pumf_dir)
kable(trouble_summary,
      caption = "Trouble sleeping distribution by cycle",
      col.names = c("Cycle", "Variable", "Scale", "Valid N", "Cat 1 %", "Cat 2 %", "Cat 3 %", "Cat 4 %", "Cat 5 %"))
```

## Summary

### Critical finding: Ontario sleep data availability

**Sleep was optional content in CCHS.** Ontario did NOT select the sleep module for most cycles.

| Cycle | Canada-wide | Ontario PUMF | Ontario 50+ N | Status |
|-------|-------------|--------------|---------------|--------|
| 2001 | Available | **Available** | ~9,900 | 3-point scale |
| 2003 | Not in PUMF | Not in PUMF | - | Excluded |
| 2005 | Not in PUMF | Not in PUMF | - | Excluded |
| 2007-2008 | Available | **NOT available** | 0 | Ontario excluded |
| 2009-2010 | Not in PUMF | Not in PUMF | - | Excluded |
| 2011-2012 | Available | **NOT available** | 0 | Ontario excluded |
| 2013-2014 | Available | **NOT available** | 0 | Ontario excluded |
| 2015-2016 | Available | **Available** | ~15,100 | 5-point scale |
| 2017-2018 | Available | **NOT available** | 0 | Ontario excluded |

### Cycles with Canada-wide sleep data

| Cycle | Hours variable | Quality variables | Scale |
|-------|---------------|-------------------|-------|
| 2001 | GENA_03 | GENA_04-06 | 3-point |
| 2007-2008 | SLP_01 | SLP_02-04 | 5-point |
| 2011-2012 | SLPG01 | SLP_02-04 | 5-point |
| 2013-2014 | SLPG01 | SLP_02-04 | 5-point |
| 2015-2016 | SLPG005 | SLP_010-020 | 5-point |
| 2017-2018 | SLPG005 | SLP_010-020 | 5-point |

### Cycles WITHOUT sleep data (PUMF)

| Cycle | Reason |
|-------|--------|
| 2003 | Sleep module not included in PUMF |
| 2005 | Sleep module not included in PUMF |
| 2009-2010 | Sleep module not included in PUMF |

### Recommendations for Ontario dementia research

**Critical limitation:** Only **2001 and 2015-2016** have Ontario sleep data in PUMF.

1. **For Ontario-specific analysis**: Only 2001 (3-point scale) and 2015-2016 (5-point scale) are usable
2. **These scales are incompatible** - cannot pool 2001 with 2015-2016 for sleep quality
3. **Hours of sleep (continuous)** may be poolable between 2001 and 2015-2016
4. **Canada-wide analysis** can use all available cycles (2001, 2007-2018 excluding 2003/2005/2009-2010)

### ICES Dictionary confirmation (v1.0.0)

**Confirmed**: Ontario Linked/Share files have the same 2003-2014 gap as PUMF.

| Period | PUMF | ONT_LINK/SHARE | Dataset |
|--------|------|----------------|---------|
| 2001 | Yes | Yes | CCHS2001_ONT_LINK_11 |
| 2002 | No | **Yes** | CCHS2002_ONT_LINK_12 |
| 2003-2014 | No | No | N/A |
| 2015-2016 | Yes | Yes | CCHS201516_ONT_SHARE |
| 2017-2018 | No | **Yes** | CCHS201718_ONT_SHARE |

**ICES advantage**: 2002 and 2017-2018 Ontario sleep data available in Linked/Share files but not PUMF.

### Scale inversion warning

- **2001**: 1 = "Most of the time" (worst), 3 = "Never" (best) - **lower = worse**
- **2007+**: 1 = "None of the time" (best), 5 = "All of the time" (worst) - **higher = worse**

**Harmonization**: Create binary "Any trouble" (Yes/No) to compare epochs.
