---
title: "CEP-004: Hearing and vision availability matrix"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  eval: true
---

## Overview

This document generates respondent count matrices for hearing and vision variables, with focus on Ontario for dementia research.

## Setup

```{r setup}
library(haven)
library(dplyr)
library(tidyr)
library(knitr)

pumf_dir <- "../../working_data_and_documentation/pumf-rdata"

# Helper function to load RData files
load_pumf_data <- function(file_path) {
  if (!file.exists(file_path)) return(NULL)
  env <- new.env()
  load(file_path, envir = env)
  # Return the first object in the environment
  get(ls(env)[1], envir = env)
}

# Age group labels (consistent across cycles for 2007+)
age_labels <- c(
  "1" = "12-14", "2" = "15-17", "3" = "18-19", "4" = "20-24",
  "5" = "25-29", "6" = "30-34", "7" = "35-39", "8" = "40-44",
  "9" = "45-49", "10" = "50-54", "11" = "55-59", "12" = "60-64",
  "13" = "65-69", "14" = "70-74", "15" = "75-79", "16" = "80+"
)

# Dementia-relevant age groups (50+)
dementia_ages <- c(10, 11, 12, 13, 14, 15, 16)
```

## Cycle configuration

```{r cycle_config}
cycles <- list(
  "2001" = list(
    file = "CCHS_2001.RData",
    prov = "GEOAGPRV",
    age = "DHHAGAGE",
    hearing = "HUIAGHER",
    vision = "HUIAGVIS"
  ),
  "2003" = list(
    file = "CCHS_2003.RData",
    prov = "GEOCGPRV",
    age = "DHHCGAGE",
    hearing = "HUICGHER",
    vision = "HUICGVIS",
    ontario_note = "NO ONTARIO DATA"
  ),
  "2009_2010" = list(
    file = "CCHS_2009_2010.RData",
    prov = "GEOGPRV",
    age = "DHHGAGE",
    hearing = "HUIGHER",
    vision = "HUIGVIS"
  ),
  "2013_2014" = list(
    file = "CCHS_2013_2014.RData",
    prov = "GEOGPRV",
    age = "DHHGAGE",
    hearing = "HUIGHER",
    vision = "HUIGVIS"
  ),
  "2017_2018" = list(
    file = "CCHS_2017_2018.RData",
    prov = "GEO_PRV",
    age = "DHHGAGE",
    hearing = "WDM_010",
    vision = "WDM_005",
    scale = "WDM"
  )
)
```

## Availability matrix: Canada-wide

```{r canada_matrix}
generate_canada_matrix <- function(cycles, pumf_dir) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) {
      results[[cycle_name]] <- data.frame(
        cycle = cycle_name,
        total_n = NA,
        hearing_n = NA,
        vision_n = NA
      )
      next
    }

    data <- load_pumf_data(file_path)
    total_n <- nrow(data)

    # Hearing
    hearing_n <- if (cfg$hearing %in% names(data)) {
      h <- data[[cfg$hearing]]
      max_val <- if (!is.null(cfg$scale) && cfg$scale == "WDM") 4 else 6
      sum(!is.na(h) & h >= 1 & h <= max_val)
    } else NA

    # Vision
    vision_n <- if (cfg$vision %in% names(data)) {
      v <- data[[cfg$vision]]
      max_val <- if (!is.null(cfg$scale) && cfg$scale == "WDM") 4 else 6
      sum(!is.na(v) & v >= 1 & v <= max_val)
    } else NA

    results[[cycle_name]] <- data.frame(
      cycle = cycle_name,
      total_n = total_n,
      hearing_n = hearing_n,
      hearing_pct = round(100 * hearing_n / total_n, 1),
      vision_n = vision_n,
      vision_pct = round(100 * vision_n / total_n, 1)
    )
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

canada_matrix <- generate_canada_matrix(cycles, pumf_dir)
kable(canada_matrix, caption = "Canada-wide availability (all ages)")
```

## Availability matrix: Ontario by age group

```{r ontario_age_matrix}
generate_ontario_age_matrix <- function(cycles, pumf_dir, age_labels, dementia_ages) {
  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) next

    data <- load_pumf_data(file_path)

    # Filter to Ontario
    if (!cfg$prov %in% names(data)) next
    ontario <- data[data[[cfg$prov]] == 35, ]

    if (nrow(ontario) == 0) {
      # No Ontario data for this cycle
      for (age_code in dementia_ages) {
        results[[paste(cycle_name, age_code)]] <- data.frame(
          cycle = cycle_name,
          age_group = age_labels[as.character(age_code)],
          ontario_n = 0,
          hearing_n = 0,
          vision_n = 0,
          note = if (!is.null(cfg$ontario_note)) cfg$ontario_note else ""
        )
      }
      next
    }

    # Analyze by age group
    for (age_code in dementia_ages) {
      age_subset <- ontario[ontario[[cfg$age]] == age_code, ]
      n <- nrow(age_subset)

      # Hearing
      hearing_n <- if (cfg$hearing %in% names(age_subset) && n > 0) {
        h <- age_subset[[cfg$hearing]]
        max_val <- if (!is.null(cfg$scale) && cfg$scale == "WDM") 4 else 6
        sum(!is.na(h) & h >= 1 & h <= max_val)
      } else 0

      # Vision
      vision_n <- if (cfg$vision %in% names(age_subset) && n > 0) {
        v <- age_subset[[cfg$vision]]
        max_val <- if (!is.null(cfg$scale) && cfg$scale == "WDM") 4 else 6
        sum(!is.na(v) & v >= 1 & v <= max_val)
      } else 0

      results[[paste(cycle_name, age_code)]] <- data.frame(
        cycle = cycle_name,
        age_group = age_labels[as.character(age_code)],
        ontario_n = n,
        hearing_n = hearing_n,
        vision_n = vision_n,
        note = if (!is.null(cfg$ontario_note)) cfg$ontario_note else ""
      )
    }
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

ontario_matrix <- generate_ontario_age_matrix(cycles, pumf_dir, age_labels, dementia_ages)
kable(ontario_matrix, caption = "Ontario availability by 5-year age group (50+)")
```

## Summary table: Age cutoffs

```{r age_cutoff_summary}
generate_cutoff_summary <- function(cycles, pumf_dir, age_labels) {
  cutoffs <- list(
    "50+" = c(10, 11, 12, 13, 14, 15, 16),
    "55+" = c(11, 12, 13, 14, 15, 16),
    "60+" = c(12, 13, 14, 15, 16)
  )

  results <- list()

  for (cycle_name in names(cycles)) {
    cfg <- cycles[[cycle_name]]
    file_path <- file.path(pumf_dir, cfg$file)

    if (!file.exists(file_path)) next

    data <- load_pumf_data(file_path)

    # Ontario only
    if (!cfg$prov %in% names(data)) next
    ontario <- data[data[[cfg$prov]] == 35, ]

    for (cutoff_name in names(cutoffs)) {
      age_codes <- cutoffs[[cutoff_name]]
      subset <- ontario[ontario[[cfg$age]] %in% age_codes, ]
      n <- nrow(subset)

      hearing_n <- if (cfg$hearing %in% names(subset) && n > 0) {
        h <- subset[[cfg$hearing]]
        max_val <- if (!is.null(cfg$scale) && cfg$scale == "WDM") 4 else 6
        sum(!is.na(h) & h >= 1 & h <= max_val)
      } else 0

      results[[paste(cycle_name, cutoff_name)]] <- data.frame(
        cycle = cycle_name,
        cutoff = cutoff_name,
        ontario_n = n,
        hearing_valid = hearing_n,
        note = if (!is.null(cfg$ontario_note)) cfg$ontario_note else ""
      )
    }
  }

  df <- do.call(rbind, results)
  rownames(df) <- NULL
  df
}

cutoff_summary <- generate_cutoff_summary(cycles, pumf_dir, age_labels)
kable(cutoff_summary, caption = "Ontario hearing availability by age cutoff")
```

## Heatmap visualization

```{r heatmap, fig.width=10, fig.height=6}
library(ggplot2)

if (exists("ontario_matrix") && nrow(ontario_matrix) > 0) {
  # Create heatmap of hearing availability
  ggplot(ontario_matrix, aes(x = cycle, y = age_group, fill = hearing_n)) +
    geom_tile(color = "white") +
    geom_text(aes(label = hearing_n), color = "black", size = 3) +
    scale_fill_gradient(low = "white", high = "steelblue",
                        name = "Valid\nrespondents") +
    labs(
      title = "Hearing variable availability: Ontario by cycle and age group",
      subtitle = "Note: 2003 has NO Ontario data for HUI hearing/vision",
      x = "CCHS Cycle",
      y = "Age Group"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Critical notes

### 2003 Ontario exclusion

The 2003 PUMF HUI grouped variables (HUICGHER, HUICGVIS) contain **NO Ontario data**. This is a data collection artifact, not a processing error. Available provinces are:

-   Atlantic provinces
-   Quebec

For Ontario dementia research, the 2003 cycle must be excluded.

### 2017-2018 Washington Group

The 2017-2018 cycle uses Washington Group measures instead of HUI:

-   WDM_005 (vision) replaces HUIGVIS
-   WDM_010 (hearing) replaces HUIGHER

The scale is 4-point (1-4) instead of 6-point (1-6). See L2_semantic_mapping.md for crosswalk.

### Missing cycles

HUI hearing/vision variables are NOT available in:

-   2005 (HUIE_07A is available but not full HUI)
-   2011-2012
-   2015-2016
