---
title: "CEP-002: Smoking Variables"
author: "Caitlin Kral, Maikol Diasparra, and Doug Manuel for the cchsflow development team"
date: "2026-01-08"
format:
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
---

::: {.callout-note collapse="true"}
## CEP metadata

| Field | Value |
|-------|-------|
| CEP | 2 |
| Title | Smoking variable harmonisation |
| Authors | Caitlin Kral, Maikol Diasparra, Doug Manuel |
| Created | 2026-01-08 |
| Status | Draft |
| Requires | CEP-001 (CSV standardization tool) |
:::

## Background

This CEP updates smoking variables in cchsflow from 35 (v2.1) to 56 variables. Updates include:

1. Extending coverage from 2015 to 2023, spanning three eras of CCHS revisions
2. Organising smoking variables into 5 subgroups: status, initiation, cessation, intensity, and pack-years
3. Adding Master file support (2001-2023)
4. Adding unified derived variables that route to the best available source across PUMF and Master

See [Harmonisation decisions](cep-002-smoking.qmd) for details on three CCHS eras, era-specific variable mappings, and PUMF vs Master differences.

## Variable count by subgroup

```{r}
#| label: tbl-variable-counts
#| echo: false
#| message: false
#| warning: false

source("../../R/table-generators.R")

generate_subgroup_summary_table(".") |>
  knitr::kable(
    col.names = c("Subgroup", "Total", "PUMF+Master", "Master-only", "Description"),
    align = c("l", "c", "c", "c", "l")
  )
```

## Recommended variables

### Primary recommendations by smoking subject

```{r}
#| label: tbl-primary-recommendations
#| echo: false
#| message: false
#| warning: false

generate_smoking_summary_table(".") |>
  knitr::kable(
    col.names = c("Subject", "Variable", "Description", "PUMF", "Master"),
    align = c("l", "l", "l", "c", "c")
  )
```

**Note**: Coverage dates are extracted from worksheet `databaseStart` fields. Variables tagged with `{recommended:primary}` are included.

::: callout-note
## PUMF 2022-2023 status variables

For 2022-2023 PUMF, use `SMKDSTY` (raw pass-through) instead of `SMKDSTY_A`. The 6-category derivation requires component variables not available in 2022+ PUMF.
:::

## All smoking variables

### 01-status (10 variables)

```{r}
#| label: tbl-status-variables
#| echo: false
#| message: false
#| warning: false

generate_variable_list_table("01-status/variables_smk_status.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

### 02-initiation (16 variables)

```{r}
#| label: tbl-initiation-variables
#| echo: false
#| message: false
#| warning: false

# Combine main initiation worksheet with unified derived variable
main_vars <- generate_variable_list_table("02-initiation/variables_smk_initiation.csv")
unified_vars <- generate_variable_list_table("02-initiation/age_start_smoking_variables.csv")
rbind(unified_vars, main_vars) |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

### 03-cessation (21 variables)

```{r}
#| label: tbl-cessation-variables
#| echo: false
#| message: false
#| warning: false

generate_variable_list_table("03-cessation/variables_smk_cessation.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

### 04-intensity (5 variables)

```{r}
#| label: tbl-intensity-variables
#| echo: false
#| message: false
#| warning: false

generate_variable_list_table("04-intensity/variables_smk_intensity.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

### 05-pack-years (2 variables)

```{r}
#| label: tbl-packyears-variables
#| echo: false
#| message: false
#| warning: false

generate_variable_list_table("05-pack-years/variables_smk_packyears.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

## Variables by use case

### Cross-cycle analysis (full 2001-2023 coverage)

For analyses requiring all CCHS cycles:

| Variable       | Description               | Notes                         |
|----------------|---------------------------|-------------------------------|
| SMKDSTY_cat3   | Current/former/never      | Recommended for most analyses |
| SMKDSTY_cat5   | 5-category status         | More detail, still comparable |
| SMK_202        | Current smoking frequency | Daily/occasional/not at all   |
| cigs_per_day   | CPD - ever-daily (unified)| Routes SMK_204/SMK_208 by status |
| pack_years_der | Pack-years                | PUMF \~15-20% error           |

### Pack-years calculation inputs

Pack-years requires these inputs:

| Component      | PUMF variable              | Master variable           |
|----------------|----------------------------|---------------------------|
| Smoking status | SMKDSTY_cat3 or SMKDSTY_A  | Same                      |
| Age started    | age_start_smoking          | Same (more precise)       |
| CPD            | cigs_per_day               | Same (but not capped)     |
| Time quit      | time_quit_smoking          | Same (but more precise)   |
| Current age    | DHHGAGE (grouped)          | DHH_AGE (continuous)      |

**Note**: The unified variables (`age_start_smoking`, `cigs_per_day`, `time_quit_smoking`) automatically route to the appropriate source variable based on smoking status.

::: callout-warning
## PUMF pack-years precision

PUMF pack-years estimates have \~15-20% relative error due to:

-   Grouped age (DHHGAGE)
-   Midpoint estimation for age started (\~±3 years)
-   CPD capped at 50
-   Midpoint estimation for time quit (\~±1.5 years)

This is acceptable for most epidemiological analyses but should be documented.
:::

### Simple smoking status

For basic current/former/never classification:

``` r
# Recommended
rec_with_table(data, "SMKDSTY_cat3")

# Returns:
# 1 = Current smoker (daily or occasional)
# 2 = Former smoker (daily or occasional)
# 3 = Never smoker
```

### Detailed smoking status

For daily/occasional distinction:

``` r
# 5-category - recommended for cross-cycle
rec_with_table(data, "SMKDSTY_cat5")

# Returns:
# 1 = Daily smoker
# 2 = Occasional smoker
# 3 = Former daily smoker
# 4 = Former occasional smoker
# 5 = Never smoker
```

## Master-only variables

These variables require RDC or ICES Master file access:

| Variable | Why Master-only       | PUMF alternative        |
|----------|-----------------------|-------------------------|
| SMK_203  | True continuous age   | SMKG203_cont (midpoint) |
| SMK_207  | True continuous age   | SMKG207_cont (midpoint) |
| SMK_040  | True continuous age   | SMKG040_cont (midpoint) |
| SMK_01C  | True continuous age   | SMKG01C_cont (midpoint) |
| SMK_01B  | True continuous count | No equivalent           |
| SMK_06C  | True quit year        | SMKG06C (categorical)   |
| SMK_09A  | True quit month       | SMK_09A_cont (midpoint) |
| SMK_09C  | True quit year        | SMKG09C_cont (midpoint) |

## Variable naming conventions

### Database suffixes

| Suffix | Meaning | Example         |
|--------|---------|-----------------|
| `_p`   | PUMF    | cchs2017_2018_p |
| `_m`   | Master  | cchs2017_2018_m |

### Variable suffixes

| Suffix | Meaning | Example |
|----------------------|-------------------------|-------------------------|
| (none) | Original CCHS | SMK_204 |
| `_cont` | Pseudo-continuous from PUMF | SMKG203_cont |
| `_cat{n}` | Re-derived categorical (n categories) | SMKDSTY_cat3, SMKDSTY_cat5 |
| `_A`, `_B` | Era-specific category sets | SMKG01C_A, SMKG01C_B |
| `_der` | Derived variable | pack_years_der |

**Note**: Do not use vanilla `_cat` without a number. Original CCHS categorical variables keep their names; `_cat{n}` is only for re-derived variables with collapsed/combined categories.

## Related documentation

-   [Harmonisation decisions](cep-002-smoking.qmd) - three CCHS eras, naming conventions, PUMF vs Master
-   [01-status](01-status.qmd) - Smoking status classification
-   [02-initiation](02-initiation.qmd) - Age started smoking
-   [03-cessation](03-cessation.qmd) - Quit timing and duration
-   [04-intensity](04-intensity.qmd) - Cigarettes per day
-   [05-pack-years](05-pack-years.qmd) - Pack-years derivation
