---
title: "01 - Smoking Status Variables"
author: "Doug Manuel, Maikol Diasparra, Caitlin St-Onge"
date: "2026-01-08"
format:
  html:
    toc: true
    toc-depth: 2
---

## Overview

| Attribute        | Value                                        |
|------------------|----------------------------------------------|
| **Subgroup**     | 01-status                                    |
| **Variables**    | 10                                           |
| **Purpose**      | Base classification for all smoking analyses |
| **Dependencies** | None                                         |
| **Dependents**   | All other smoking subgroups                  |

Smoking status variables provide the foundation for smoking research. They classify respondents into smoking categories (current, former, never) with varying levels of detail.

## Variables

```{r}
#| label: tbl-status-variables
#| echo: false
#| message: false
#| warning: false

source("../../R/table-generators.R")

generate_variable_list_table("01-status/variables_smk_status.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

## PUMF cycle coverage

```{r}
#| label: tbl-status-coverage
#| echo: false
#| message: false
#| warning: false

generate_subject_coverage_matrix("01-status") |>
  knitr::kable(align = c("l", rep("c", 13)))
```

## Key decisions

### SMKDSTY semantic break in 2015

The StatsCan derived variable SMKDSTY changed category definitions in 2015:

| Category | Pre-2015 meaning         | 2015+ meaning       |
|----------|--------------------------|---------------------|
| 5        | Former occasional smoker | Experimental smoker |

**Solution**: Created harmonized variables that avoid this break:

-   `SMKDSTY_cat3` - Collapses to current/former/never
-   `SMKDSTY_cat5` - Merges problematic categories
-   `SMKDSTY_A` - Pre-2015 structure for all cycles (derives 2015+ from components)

### All-cycle harmonized variables

| Use case | Variable | Why |
|---------------------------|---------------------------|------------------|
| Simple status | SMKDSTY_cat3 | Full coverage, no semantic breaks |
| Detailed status | SMKDSTY_cat5 | Daily/occasional distinction, comparable |
| Current frequency | SMK_202 | Direct question, full coverage |
| Pack-years input | SMKDSTY_A | 6-category needed for pack_years_der derivation |

### 2022+ PUMF coverage

SMK_01A and SMK_05D are not in 2022-2023 PUMF (module split). Users should:

-   Use SMKDSTY-based variables for 2022-2023 PUMF
-   Master files have full coverage via CSS_15 and SPU_05

## Era mappings

| Variable | Era 1 (2001-2014)     | Era 2 (2015-2021) | Era 3 (2022+) |
|----------|-----------------------|-------------------|---------------|
| SMK_202  | SMKA/C/E_202, SMK_202 | SMK_005           | CSS_05        |
| SMK_01A  | SMKA/C/E_01A, SMK_01A | SMK_020           | CSS_15        |
| SMK_05D  | SMKA/C/E_05D, SMK_05D | SMK_030           | SPU_05        |
| SMKDSTY  | SMKA/C/EDSTY, SMKDSTY | SMKDVSTY          | SMKDVSTY      |

## Validation status

| Level | Status  | Notes                                  |
|-------|---------|----------------------------------------|
| L0    | ✓       | Schema compliant                       |
| L1    | ✓       | Database names validated               |
| L2    | ✓       | Source references verified against DDI |
| L3    | ✓       | Category codes verified                |
| L4    | ✓       | Semantic consistency reviewed          |
| L5    | ✓       | Integration tests (below)              |

## Integration testing

This section validates the primary recommended variable `SMKDSTY_cat6` using actual PUMF data. Testing includes:

1. **Ground truth** (pre-2015 cycles): Direct comparison where source→target is 1:1 mapping
2. **rec_with_table()**: Validation that harmonization produces expected results
3. **Clinical reasonableness**: Distribution checks against epidemiological expectations

### Clinical expectations

| Category | Label | Expected prevalence |
|----------|-------|---------------------|
| 1 | Daily smoker | 10-25% (declining trend) |
| 2 | Occasional (former daily) | 2-4% |
| 3 | Always occasional | 3-5% |
| 4 | Former daily | 25-35% |
| 5 | Former occasional | 5-10% |
| 6 | Never smoked | 30-40% |

**Trend expectation**: Daily smoking prevalence should decline over time (public health success).

```{r}
#| label: setup-integration
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)
library(here)

# Project paths
project_root <- here::here()
rdata_dir <- file.path(project_root, "working_data_and_documentation/pumf-rdata")

# Source cchsflow functions
original_wd <- getwd()
setwd(project_root)
source("R/strings.R")
source("R/recode-with-table.R")
setwd(original_wd)

# Load worksheets from this subgroup
variables_csv <- file.path(project_root, "ceps/cep-002-smoking/01-status/variables_smk_status.csv")
variable_details_csv <- file.path(project_root, "ceps/cep-002-smoking/01-status/variable_details_smk_status.csv")

variables <- read.csv(variables_csv, stringsAsFactors = FALSE)
variable_details <- read.csv(variable_details_csv, stringsAsFactors = FALSE)

# Cycle order
cycle_order <- c(
  "2001", "2003", "2005", "2007_2008", "2009_2010",
  "2011_2012", "2013_2014", "2015_2016", "2017_2018",
  "2019_2020", "2022"
)
```

### Ground truth: SMKDSTY_cat6 (pre-2015 cycles)

For cycles 2001-2014, `SMKDSTY_cat6` is a direct 1:1 pass-through from the source variable (SMKADSTY, SMKCDSTY, SMKEDSTY, or SMKDSTY). This allows ground truth validation.

```{r}
#| label: ground-truth-status
#| code-fold: true

# Derive ground truth from raw PUMF data (pre-2015 only)
derive_smkdsty_ground_truth <- function(df, cycle) {

  # Identify source variable by cycle
  if (cycle == "2001") {
    var_name <- "SMKADSTY"
  } else if (cycle == "2003") {
    var_name <- "SMKCDSTY"
  } else if (cycle == "2005") {
    var_name <- "SMKEDSTY"
  } else if (cycle %in% c("2007_2008", "2009_2010", "2011_2012", "2013_2014")) {
    var_name <- "SMKDSTY"
  } else {
    return(list(values = rep(NA, nrow(df)), source_var = "N/A (derived)",
                method = "2015+ requires derivation"))
  }

  if (var_name %in% names(df)) {
    return(list(values = df[[var_name]], source_var = var_name,
                method = "Direct pass-through"))
  } else {
    return(list(values = rep(NA, nrow(df)), source_var = paste(var_name, "(missing)"),
                method = "Variable not in dataset"))
  }
}

# Process all pre-2015 cycles for ground truth
ground_truth_results <- list()
pre2015_cycles <- c("2001", "2003", "2005", "2007_2008", "2009_2010",
                     "2011_2012", "2013_2014")

for (cycle in pre2015_cycles) {
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) {
    message("Skipping ", cycle, " - file not found")
    next
  }

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  derived <- derive_smkdsty_ground_truth(df, cycle)
  values <- derived$values

  # Count by category (valid = 1-6)
  n_total <- nrow(df)
  cat_counts <- table(factor(values, levels = 1:6))
  n_valid <- sum(cat_counts)

  ground_truth_results[[cycle]] <- data.frame(
    Cycle = cycle,
    N = n_total,
    Source = derived$source_var,
    Daily = as.numeric(cat_counts[1]),
    Occ_fmr = as.numeric(cat_counts[2]),
    Always_occ = as.numeric(cat_counts[3]),
    Fmr_daily = as.numeric(cat_counts[4]),
    Fmr_occ = as.numeric(cat_counts[5]),
    Never = as.numeric(cat_counts[6]),
    Pct_daily = round(100 * cat_counts[1] / n_valid, 1),
    Pct_never = round(100 * cat_counts[6] / n_valid, 1)
  )
}

ground_truth_df <- do.call(rbind, ground_truth_results)
rownames(ground_truth_df) <- NULL
```

```{r}
#| label: tbl-ground-truth
#| tbl-cap: "Ground truth: SMKDSTY distribution from raw PUMF (pre-2015 cycles)"

if (!is.null(ground_truth_df) && nrow(ground_truth_df) > 0) {
  kable(ground_truth_df[, c("Cycle", "N", "Source", "Daily", "Fmr_daily", "Never",
                             "Pct_daily", "Pct_never")],
        col.names = c("Cycle", "N", "Source", "Daily", "Fmr Daily", "Never",
                      "% Daily", "% Never"),
        format.args = list(big.mark = ","))
} else {
  cat("No ground truth data available. Check that PUMF data files exist.")
}
```

### rec_with_table() validation

Test that `rec_with_table()` produces the same results as ground truth for pre-2015 cycles, and produces valid results for 2015+ cycles.

```{r}
#| label: rec-with-table-test
#| code-fold: true

# Test rec_with_table() against ground truth
rwt_results <- list()
test_results <- list()

for (cycle in cycle_order) {
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) next

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  # Database name for this cycle
  db_name <- paste0("cchs", cycle, "_p")

  # Apply rec_with_table for SMKDSTY_cat6
  tryCatch({
    harmonized <- rec_with_table(
      data = df,
      variables = variables,
      variable_details = variable_details,
      database_name = db_name,
      var_names = "SMKDSTY_cat6",
      log = FALSE
    )

    if ("SMKDSTY_cat6" %in% names(harmonized)) {
      rwt_values <- harmonized$SMKDSTY_cat6
      cat_counts <- table(factor(rwt_values, levels = 1:6))
      n_valid <- sum(cat_counts)

      rwt_results[[cycle]] <- data.frame(
        Cycle = cycle,
        N = nrow(df),
        Method = "rec_with_table",
        Daily = as.numeric(cat_counts[1]),
        Fmr_daily = as.numeric(cat_counts[4]),
        Never = as.numeric(cat_counts[6]),
        Pct_daily = round(100 * cat_counts[1] / n_valid, 1),
        Pct_never = round(100 * cat_counts[6] / n_valid, 1)
      )

      # Compare with ground truth for pre-2015 cycles
      if (cycle %in% pre2015_cycles && cycle %in% names(ground_truth_results)) {
        gt_daily <- ground_truth_results[[cycle]]$Pct_daily
        rwt_daily <- rwt_results[[cycle]]$Pct_daily
        match_status <- ifelse(abs(gt_daily - rwt_daily) < 0.1, "PASS", "FAIL")
        test_results[[cycle]] <- list(
          cycle = cycle,
          ground_truth = gt_daily,
          rec_with_table = rwt_daily,
          match = match_status
        )
      }
    }
  }, error = function(e) {
    message("rec_with_table failed for ", cycle, ": ", e$message)
  })
}

rwt_df <- do.call(rbind, rwt_results)
rownames(rwt_df) <- NULL
```

```{r}
#| label: tbl-rec-with-table
#| tbl-cap: "rec_with_table() results: SMKDSTY_cat6 across all cycles"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  kable(rwt_df,
        col.names = c("Cycle", "N", "Method", "Daily", "Fmr Daily", "Never",
                      "% Daily", "% Never"),
        format.args = list(big.mark = ","))
} else {
  cat("No rec_with_table results available. Check that PUMF data files exist.")
}
```

### Ground truth comparison

```{r}
#| label: tbl-comparison
#| tbl-cap: "Ground truth vs rec_with_table() comparison (pre-2015 cycles)"

if (length(test_results) > 0) {
  comparison_df <- do.call(rbind, lapply(test_results, as.data.frame))
  rownames(comparison_df) <- NULL

  kable(comparison_df,
        col.names = c("Cycle", "Ground Truth (% Daily)", "rec_with_table (% Daily)", "Match"))
} else {
  cat("No comparison data available")
}
```

### Clinical assessment

```{r}
#| label: clinical-assessment
#| code-fold: true

# Assess clinical reasonableness
assessment <- list()

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  # Check 1: Daily smoking prevalence in expected range
  daily_range <- range(rwt_df$Pct_daily, na.rm = TRUE)
  check_daily <- daily_range[1] >= 5 && daily_range[2] <= 30
  assessment$daily_range <- list(
    check = "Daily smoking 5-30%",
    observed = paste0(daily_range[1], "% - ", daily_range[2], "%"),
    pass = check_daily
  )

  # Check 2: Never smokers in expected range
  never_range <- range(rwt_df$Pct_never, na.rm = TRUE)
  check_never <- never_range[1] >= 25 && never_range[2] <= 50
  assessment$never_range <- list(
    check = "Never smokers 25-50%",
    observed = paste0(never_range[1], "% - ", never_range[2], "%"),
    pass = check_never
  )

  # Check 3: Declining daily smoking trend
  early_cycles <- rwt_df$Pct_daily[1:3]
  late_cycles <- rwt_df$Pct_daily[(nrow(rwt_df)-2):nrow(rwt_df)]
  declining_trend <- mean(early_cycles, na.rm = TRUE) > mean(late_cycles, na.rm = TRUE)
  assessment$trend <- list(
    check = "Daily smoking declining over time",
    observed = paste0("Early avg: ", round(mean(early_cycles, na.rm = TRUE), 1),
                      "%, Late avg: ", round(mean(late_cycles, na.rm = TRUE), 1), "%"),
    pass = declining_trend
  )
}

# Display assessment
assessment_df <- do.call(rbind, lapply(assessment, function(x) {
  data.frame(Check = x$check, Observed = x$observed,
             Result = ifelse(x$pass, "✓ PASS", "✗ FAIL"))
}))
```

```{r}
#| label: tbl-assessment
#| tbl-cap: "Clinical reasonableness assessment"

if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) {
  kable(assessment_df, row.names = FALSE)
} else {
  cat("No clinical assessment available - no data to assess.")
}
```

### Daily smoking trend

```{r}
#| label: fig-daily-trend
#| fig-cap: "Daily smoker prevalence trend (ground truth = rec_with_table for pre-2015)"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  # Extract numeric year for plotting
  years <- as.numeric(gsub("_.*", "", rwt_df$Cycle))

  plot(years, rwt_df$Pct_daily, type = "b", pch = 19, col = "darkred",
       xlab = "CCHS Year", ylab = "% Daily Smokers",
       main = "Daily Smoker Prevalence Trend",
       ylim = c(0, 25))

  # Add reference line at 2015 (semantic break)
  abline(v = 2015, lty = 2, col = "gray50")
  text(2015, 24, "2015 semantic break", pos = 4, cex = 0.8, col = "gray50")
}
```

### Test summary

```{r}
#| label: test-summary

# Summarize all tests
n_gt_tests <- length(test_results)
n_gt_pass <- if (n_gt_tests > 0) sum(sapply(test_results, function(x) x$match == "PASS")) else 0

n_clinical <- if (exists("assessment_df") && !is.null(assessment_df)) nrow(assessment_df) else 0
n_clinical_pass <- if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) sum(grepl("PASS", assessment_df$Result)) else 0

n_cycles <- if (!is.null(rwt_df)) nrow(rwt_df) else 0

cat("## SMKDSTY_cat6 Integration Test Summary\n\n")
cat("**Ground truth tests (pre-2015):**", n_gt_pass, "/", n_gt_tests, "passed\n")
cat("**Clinical reasonableness:**", n_clinical_pass, "/", n_clinical, "passed\n")
cat("**Cycles tested:**", n_cycles, "\n")

if (n_cycles == 0) {
  cat("\n⚠ **No PUMF data available for testing**\n")
} else if (n_gt_pass == n_gt_tests && n_clinical_pass == n_clinical) {
  cat("\n✓ **L5 INTEGRATION TESTS: PASSED**\n")
} else {
  cat("\n✗ **L5 INTEGRATION TESTS: ISSUES DETECTED**\n")
}
```

## Appendix

Detailed working documentation:

-   [L0: Documentation assessment](01-status/L0_documentation_assessment.md)
-   [L1: Variable concordance](01-status/L1_variable_concordance.md)
-   [L2: Semantic mapping](01-status/L2_semantic_mapping.md)
-   [L3: Worksheet draft](01-status/L3_worksheet_draft.md)
-   [L4: DV specifications](01-status/L4_dv_specifications.md)
-   [variables_smk_status.csv](01-status/variables_smk_status.csv)
-   [variable_details_smk_status.csv](01-status/variable_details_smk_status.csv)