---
title: "03 - Smoking Cessation Variables"
author: "Doug Manuel, Maikol Diasparra, Caitlin St-Onge"
date: "2026-01-08"
format:
  html:
    toc: true
    toc-depth: 2
---

## Overview

| Attribute | Value |
|-----------|-------|
| **Subgroup** | 03-cessation |
| **Variables** | 23 (22 base + 1 unified) |
| **Purpose** | Quit timing and duration (for pack-years calculation) |
| **Dependencies** | 01-status, ADM_MOI_I, ADM_YOI_I (for precise timing) |
| **Dependents** | 05-pack-years |

Cessation variables capture when and how respondents quit smoking. These are critical for calculating smoking duration and pack-years exposure.

## Variables

```{r}
#| label: tbl-cessation-variables
#| echo: false
#| message: false
#| warning: false

source("../../R/table-generators.R")

generate_variable_list_table("03-cessation/variables_smk_cessation.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

## PUMF cycle coverage

```{r}
#| label: tbl-cessation-coverage
#| echo: false
#| message: false
#| warning: false

generate_subject_coverage_matrix("03-cessation") |>
  knitr::kable(align = c("l", rep("c", 13)))
```

## Key decisions

### Time since quit precision

| Source | PUMF precision | Master precision |
|--------|----------------|------------------|
| SMK_06A | ~±2.5 years (midpoint) | - |
| SMK_06C | - | Exact year |
| Derived | ~±1.5 years | Month/year exact |

### Midpoint values for SMK_06A_cont

| Category | Years range | Midpoint |
|----------|-------------|----------|
| 1 | <1 year | 0.5 |
| 2 | 1-<2 years | 1.5 |
| 3 | 2-<3 years | 2.5 |
| 4 | 3-<5 years | 4 |
| 5 | 5-<10 years | 7.5 |
| 6 | 10+ years | 15 |

### Derived variable logic

::: callout-important
## Recommended: Use `time_quit_complete` for pack-years

**`time_quit_complete`** correctly handles the SMK_10_gate logic for "reducers" (former daily smokers who continued occasional before quitting completely). Use this for pack-years calculation.

**`time_quit_smoking`** (secondary) does not handle SMK_10_gate and gives incorrect values for reducers. It remains available for backward compatibility and for 2001 data (where SMK_10_gate was not collected).
:::

#### `time_quit_complete` (recommended)

Calculated from `DerivedVar::[SMK_10_gate, SMK_09A_cont, SMK_10A_cont, SMK_06A_cont]`

```
For former daily smokers:
  - If SMK_10_gate = 1 (quit when stopped daily): use SMK_09A_cont
  - If SMK_10_gate = 2 (continued occasional then quit): use SMK_10A_cont

For former occasional (never daily): use SMK_06A_cont

For current smokers: NA (not applicable)
For never smokers: NA (not applicable)
```

**Availability:** 2003+ only (SMK_10_gate not collected in 2001)

#### `time_quit_smoking` (secondary)

Calculated from `DerivedVar::[SMK_09A_cont, SMK_06A_cont, SPU_25I]`

```
Priority order for all former smokers:
  1. SMK_09A_cont (former daily smokers)
  2. SMK_06A_cont (former occasional, never daily)
  3. SPU_25I (2022+ Master direct value)

For current smokers: NA (not applicable)
For never smokers: NA (not applicable)
```

**Limitation:** Uses SMK_09A for all former daily smokers regardless of SMK_10_gate. This gives incorrect values for "reducers" (SMK_10=2) who continued occasional smoking before quitting.

**Use case:** 2001 data only, or when backward compatibility is required.

#### `quit_pathway`

Calculated from `DerivedVar::[SMKDSTY_cat5, SMK_10_gate]`

```
1 = direct quit (quit when stopped daily)
2 = gradual quit (continued occasional then quit)
3 = former occasional (never daily)
```

### Interview date dependency

Precise cessation timing requires interview date variables to convert "quit month/year" responses to "years since quit":

| Variable | Description | Cycles |
|----------|-------------|--------|
| ADM_MOI_I | Interview month (1-12) | 2001-2023 (Master) |
| ADM_YOI_I | Interview year | 2001-2023 (Master) |

The continuous cessation variables (SMK_06A_cont, SMK_09A_cont, SMK_10A_cont) use interview date when:

1. **Category 1 (<1 year)**: Calculate precise time from quit month (SMK_xxB_I) and interview month
2. **Category 4 (3+ years)**: Calculate precise time from quit year (SMK_xxC_I) and interview year
3. **2022 cycle**: Calculate from SPU_xxA_I (month) and SPU_xxB_I (year) relative to interview date

```
years_since_quit = ADM_YOI_I - quit_year + (ADM_MOI_I - quit_month) / 12
```

For PUMF files where interview date is not available, midpoint imputation is used instead.

### Universe complexity

Cessation variables have complex skip patterns based on smoking history:

| Variable series | Universe |
|-----------------|----------|
| SMK_06A, SMK_06C, SMKG06C | Former occasional smokers (never daily) |
| SMK_09A, SMK_09C, SMKG09C | Former daily smokers |
| SMK_10A, SMK_10_gate | Former daily who continued occasional then quit |
| SMKDVSTP, SMKDGSTP | Former smokers who quit completely |

## Era mappings

Based on worksheet `variableStart` mappings:

| Concept | Era 1 (2001-2014) | Era 2 (2015-2021) | Era 3 (2022+) |
|---------|-------------------|-------------------|---------------|
| Quit time (occ) | SMKA_06A, SMK_06A | SMK_060 | SPU_10 |
| Quit time (daily) | SMKA_09A, SMK_09A | SMK_080 | SPU_25 |
| Quit time (reducer) | SMK_10A | SMK_100 | SPU_35 |
| Quit gate | SMK_10 | SMK_095 | SPU_30 |
| Quit year (occ) | SMK_06C | - | - |
| Quit year (daily) | - | - | SPU_25I |

## Valid ranges

### Time-based ranges

| Variable | Valid range | Rationale |
|----------|-------------|-----------|
| SMK_06A_cont | 0-80 | Years since quit; max = age - 5 |
| SMK_09A_cont | 0-11 | Months since quit (within past year) |
| time_quit_smoking | 0-80 | Derived years; cannot exceed age minus ~5 |
| years_smoked | 0-80 | Total duration; capped by lifespan |
| age_quit | 5-100 | Must be > age started, ≤ current age |

### Cross-field constraints

| Rule | Constraint | Action if violated |
|------|------------|-------------------|
| Age quit ≥ age started | age_quit ≥ SMK_203/207 | Set to NA(b) |
| Age quit ≤ current age | age_quit ≤ DHH_AGE | Set to NA(b) |
| Years since quit + age started ≤ current age | Plausibility check | Flag for review |
| Years smoked ≤ current age - 5 | smoking_years ≤ DHH_AGE - 5 | Cap at maximum |

### Derived variable plausibility

`time_quit_smoking` calculation should satisfy:

```
time_quit_smoking = current_age - age_quit
                  = current_age - (age_started + years_smoked)
```

If these don't reconcile (within midpoint estimation error), flag for review.

### Edge cases

| Situation | Assessment | Action |
|-----------|------------|--------|
| Quit <1 year ago | Valid (category 1) | Use midpoint 0.5 years |
| Quit 10+ years ago | Valid (category 6) | Use midpoint 15 years or cap at age-based max |
| Time since quit > age - 5 | Impossible | Set to NA(b) |
| Age quit = age started | Implausible but possible | Flag for review |

## Validation status

| Level | Status | Notes |
|-------|--------|-------|
| L0 | ✓ | Schema compliant |
| L1 | ✓ | Database names validated |
| L2 | ✓ | Source references verified |
| L3 | ✓ | Category codes verified |
| L4 | ✓ | Universe logic reviewed |
| L5 | ✓ | Integration tests (below) |

## Integration testing

This section validates the primary recommended variable `time_quit_smoking` using actual PUMF data. Testing includes:

1. **rec_with_table()**: Validation that the derived variable produces expected results
2. **Universe validation**: Only former smokers (SMKDSTY 4, 5) should have values
3. **Clinical reasonableness**: Distribution checks against epidemiological expectations

**Note**: `time_quit_smoking` is a derived variable. Ground truth comparison is not possible in the same way as direct pass-through variables.

### Clinical expectations

| Metric | Expected value |
|--------|----------------|
| Mean time quit | 10-20 years (former smokers quit across lifespan) |
| Range | 0-80 years (per worksheet bounds) |
| Distribution | Right-skewed (more recent quitters in survey) |
| Universe | Former smokers only (SMKDSTY 4, 5) |

**Epidemiological context**: Former smokers have typically quit between 1 and 30 years ago, with a long tail of long-term quitters.

```{r}
#| label: setup-integration-cessation
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)
library(here)

# Project paths
project_root <- here::here()
rdata_dir <- file.path(project_root, "working_data_and_documentation/pumf-rdata")

# Source cchsflow functions
original_wd <- getwd()
setwd(project_root)
source("R/strings.R")
source("R/recode-with-table.R")
setwd(original_wd)

# Load cessation worksheets
cessation_variables <- read.csv(file.path(project_root, "ceps/cep-002-smoking/03-cessation/variables_smk_cessation.csv"), stringsAsFactors = FALSE)
cessation_variable_details <- read.csv(file.path(project_root, "ceps/cep-002-smoking/03-cessation/variable_details_smk_cessation.csv"), stringsAsFactors = FALSE)

# Also need status variables for SMKDSTY_cat6
status_variables <- read.csv(file.path(project_root, "ceps/cep-002-smoking/01-status/variables_smk_status.csv"), stringsAsFactors = FALSE)
status_variable_details <- read.csv(file.path(project_root, "ceps/cep-002-smoking/01-status/variable_details_smk_status.csv"), stringsAsFactors = FALSE)

all_variables <- rbind(cessation_variables, status_variables)
all_variable_details <- rbind(cessation_variable_details, status_variable_details)

# Cycle order
cycle_order <- c(
  "2001", "2003", "2005", "2007_2008", "2009_2010",
  "2011_2012", "2013_2014", "2015_2016", "2017_2018",
  "2019_2020"
)
```

### rec_with_table() validation

Test that `time_quit_smoking` produces continuous values in the expected range for former smokers.

```{r}
#| label: rec-with-table-cessation
#| code-fold: true

# Test rec_with_table() for time_quit_smoking across cycles
rwt_results <- list()

for (cycle in cycle_order) {
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) next

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  db_name <- paste0("cchs", cycle, "_p")

  tryCatch({
    harmonized <- rec_with_table(
      data = df,
      variables = all_variables,
      variable_details = all_variable_details,
      database_name = db_name,
      var_names = c("time_quit_smoking", "SMKDSTY_cat6"),
      log = FALSE
    )

    if ("time_quit_smoking" %in% names(harmonized)) {
      time_values <- harmonized$time_quit_smoking
      status_values <- harmonized$SMKDSTY_cat6

      # Filter to former smokers (SMKDSTY 4, 5)
      former <- status_values %in% c(4, 5)
      time_former <- time_values[former]
      time_valid <- time_former[!is.na(time_former)]

      # Calculate statistics
      if (length(time_valid) > 0) {
        rwt_results[[cycle]] <- data.frame(
          Cycle = cycle,
          N = nrow(df),
          N_former = sum(former, na.rm = TRUE),
          N_valid = length(time_valid),
          Mean = round(mean(time_valid), 1),
          Median = round(median(time_valid), 1),
          Min = round(min(time_valid), 1),
          Max = round(max(time_valid), 1),
          Pct_valid = round(100 * length(time_valid) / sum(former, na.rm = TRUE), 1),
          check.names = FALSE
        )
      }
    }
  }, error = function(e) {
    message("rec_with_table failed for ", cycle, ": ", e$message)
  })
}

rwt_df <- do.call(rbind, rwt_results)
rownames(rwt_df) <- NULL
```

```{r}
#| label: tbl-cessation-results
#| tbl-cap: "rec_with_table() results: time_quit_smoking across PUMF cycles"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  kable(rwt_df,
        col.names = c("Cycle", "N Total", "Former", "Valid", "Mean", "Median", "Min", "Max", "% Valid"),
        format.args = list(big.mark = ","))
}
```

### Universe validation

Verify that `time_quit_smoking` is NA for current smokers and never smokers.

```{r}
#| label: universe-validation-cessation
#| code-fold: true

# Check that non-former smokers have NA
universe_check <- list()

for (cycle in c("2007_2008", "2015_2016")) {  # Sample cycles from each era
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) next

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  db_name <- paste0("cchs", cycle, "_p")

  tryCatch({
    harmonized <- rec_with_table(
      data = df,
      variables = all_variables,
      variable_details = all_variable_details,
      database_name = db_name,
      var_names = c("time_quit_smoking", "SMKDSTY_cat6"),
      log = FALSE
    )

    status <- harmonized$SMKDSTY_cat6
    time_quit <- harmonized$time_quit_smoking

    # Count by status
    for (s in 1:6) {
      in_status <- status == s & !is.na(status)
      n_in_status <- sum(in_status, na.rm = TRUE)
      n_has_time <- sum(!is.na(time_quit[in_status]), na.rm = TRUE)
      pct_has_time <- if (n_in_status > 0) round(100 * n_has_time / n_in_status, 1) else NA

      universe_check[[paste0(cycle, "_", s)]] <- data.frame(
        Cycle = cycle,
        Status = s,
        Status_label = c("Daily", "Occ (fmr daily)", "Always occ", "Fmr daily", "Fmr occ", "Never")[s],
        N_status = n_in_status,
        N_has_time = n_has_time,
        Pct_has_time = pct_has_time,
        Expected = ifelse(s %in% c(4, 5), "Has time", "NA"),
        check.names = FALSE
      )
    }
  }, error = function(e) {
    message("Universe check failed for ", cycle, ": ", e$message)
  })
}

universe_df <- do.call(rbind, universe_check)
rownames(universe_df) <- NULL
```

```{r}
#| label: tbl-universe-cessation
#| tbl-cap: "Universe validation: time_quit_smoking by smoking status"

if (!is.null(universe_df) && nrow(universe_df) > 0) {
  kable(universe_df,
        col.names = c("Cycle", "Status", "Label", "N", "Has Time", "% Has Time", "Expected"))
}
```

### Clinical assessment

```{r}
#| label: clinical-assessment-cessation
#| code-fold: true

assessment <- list()

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  # Check 1: Mean time quit in expected range
  mean_range <- range(rwt_df$Mean, na.rm = TRUE)
  check_mean <- mean_range[1] >= 5 && mean_range[2] <= 30
  assessment$mean_range <- list(
    check = "Mean time quit 5-30 years",
    observed = paste0(mean_range[1], " - ", mean_range[2], " years"),
    pass = check_mean
  )

  # Check 2: Valid percentage reasonable (former smokers may have missing)
  pct_valid_range <- range(rwt_df$Pct_valid, na.rm = TRUE)
  check_coverage <- pct_valid_range[1] >= 50  # Lower threshold - cessation often has more missing
  assessment$coverage <- list(
    check = "Coverage ≥50% of former smokers",
    observed = paste0(pct_valid_range[1], "% - ", pct_valid_range[2], "%"),
    pass = check_coverage
  )

  # Check 3: Universe correctness (current/never should have low %)
  if (exists("universe_df") && nrow(universe_df) > 0) {
    non_former <- universe_df[universe_df$Status %in% c(1, 2, 3, 6), ]
    max_pct_non_former <- max(non_former$Pct_has_time, na.rm = TRUE)
    check_universe <- max_pct_non_former < 5
    assessment$universe <- list(
      check = "Current/never smokers have <5% with time_quit",
      observed = paste0("Max: ", max_pct_non_former, "%"),
      pass = check_universe
    )
  }

  # Check 4: Values within bounds (0-80)
  max_value <- max(rwt_df$Max, na.rm = TRUE)
  min_value <- min(rwt_df$Min, na.rm = TRUE)
  check_bounds <- min_value >= 0 && max_value <= 80
  assessment$bounds <- list(
    check = "Values within [0, 80] years",
    observed = paste0(min_value, " - ", max_value, " years"),
    pass = check_bounds
  )
}

assessment_df <- do.call(rbind, lapply(assessment, function(x) {
  data.frame(Check = x$check, Observed = x$observed,
             Result = ifelse(x$pass, "✓ PASS", "✗ FAIL"))
}))
```

```{r}
#| label: tbl-assessment-cessation
#| tbl-cap: "Clinical reasonableness assessment"

if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) {
  kable(assessment_df, row.names = FALSE)
}
```

### Time quit distribution

```{r}
#| label: fig-time-quit-distribution
#| fig-cap: "Mean time since quit smoking by cycle"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  years <- as.numeric(gsub("_.*", "", rwt_df$Cycle))

  plot(years, rwt_df$Mean, type = "b", pch = 19, col = "steelblue",
       xlab = "CCHS Year", ylab = "Mean Years Since Quit",
       main = "Time Since Quit Smoking",
       ylim = c(0, max(rwt_df$Mean, na.rm = TRUE) * 1.2))

  lines(years, rwt_df$Median, type = "b", pch = 17, col = "darkgreen", lty = 2)

  legend("topright", legend = c("Mean", "Median"),
         col = c("steelblue", "darkgreen"), pch = c(19, 17), lty = c(1, 2))
}
```

### Test summary

```{r}
#| label: test-summary-cessation

n_clinical <- if (exists("assessment_df") && !is.null(assessment_df)) nrow(assessment_df) else 0
n_clinical_pass <- if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) sum(grepl("PASS", assessment_df$Result)) else 0
n_cycles <- if (!is.null(rwt_df)) nrow(rwt_df) else 0

cat("## time_quit_smoking Integration Test Summary\n\n")
cat("**Clinical reasonableness:**", n_clinical_pass, "/", n_clinical, "passed\n")
cat("**Cycles tested:**", n_cycles, "\n")

if (n_cycles == 0) {
  cat("\n⚠ **No PUMF data available for testing**\n")
} else if (n_clinical == 0 || n_clinical_pass == n_clinical) {
  cat("\n✓ **L5 INTEGRATION TESTS: PASSED**\n")
} else {
  cat("\n✗ **L5 INTEGRATION TESTS: ISSUES DETECTED**\n")
}
```

## Appendix

### Authoritative worksheets

- [variables_smk_cessation.csv](03-cessation/variables_smk_cessation.csv) - Variable definitions (22 base variables)
- [variable_details_smk_cessation.csv](03-cessation/variable_details_smk_cessation.csv) - Recoding rules
- [time_quit_complete_variables.csv](03-cessation/time_quit_complete_variables.csv) - Unified quit timing variable (new)
- [time_quit_complete_variable_details.csv](03-cessation/time_quit_complete_variable_details.csv) - Unified quit timing details (new)

### Supporting documentation

- [L0: Documentation assessment](03-cessation/L0-assessment-smoking-cessation.md)
- [L1: Variable concordance](03-cessation/L1-variable-concordance.md)
- [L2: Semantic mapping](03-cessation/L2-semantic-mapping.md)
- [L3: Worksheet draft](03-cessation/L3_worksheet_draft.md)
- [L4: DV specifications](03-cessation/L4_dv_specifications.md)
