---
title: "04 - Smoking Intensity Variables"
author: "Doug Manuel, Maikol Diasparra, Caitlin St-Onge"
date: "2026-01-08"
format:
  html:
    toc: true
    toc-depth: 2
---

## Overview

| Attribute | Value |
|-----------|-------|
| **Subgroup** | 04-intensity |
| **Variables** | 5 |
| **Purpose** | Cigarettes per day (for pack-years calculation) |
| **Dependencies** | 01-status |
| **Dependents** | 05-pack-years |

Intensity variables capture how much respondents smoke or smoked. These provide the "cigarettes per day" component of pack-years calculation.

## Variables

```{r}
#| label: tbl-intensity-variables
#| echo: false
#| message: false
#| warning: false

source("../../R/table-generators.R")

generate_variable_list_table("04-intensity/variables_smk_intensity.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

## PUMF cycle coverage

```{r}
#| label: tbl-intensity-coverage
#| echo: false
#| message: false
#| warning: false

generate_subject_coverage_matrix("04-intensity") |>
  knitr::kable(align = c("l", rep("c", 13)))
```

::: {.callout-tip}
## Full coverage

All 5 intensity variables have complete PUMF and Master coverage across all 13 CCHS cycles (2001-2023). This is unusual among smoking variables.
:::

## Key decisions

### PUMF capping

| Variable | PUMF range | Master range |
|----------|------------|--------------|
| SMK_204 | 1-50 (capped) | 1-99 (true) |
| SMK_208 | 1-50 (capped) | 1-99 (true) |

**Impact on pack-years**: Heavy smokers (>50 CPD) are underestimated in PUMF. This affects ~2-3% of respondents but can significantly impact pack-years for heavy smokers.

### Occasional smoker intensity

For occasional smokers, pack-years uses:

```
CPD_equivalent = (SMK_05B × SMK_05C) / 30
```

This converts "cigarettes on days smoked" × "days per month" to a daily average.

### Unified daily intensity: cigs_per_day

The `cigs_per_day` variable combines SMK_204 (current daily) and SMK_208 (former daily) into a single derived variable representing cigarettes per day when smoking daily, for anyone who ever smoked daily.

**Rationale**: SMK_204 and SMK_208 are mutually exclusive by design and measure the same concept (daily smoking intensity). The unified variable simplifies the mental model and aligns with how `age_start_smoking` and `time_quit_smoking` work.

**Routing logic**:

| SMKDSTY_A status | Source variable | Result |
|------------------|-----------------|--------|
| 1 (current daily) | SMK_204 | Cigarettes per day |
| 2 (occasional, former daily) | SMK_208 | Cigarettes per day (when smoked daily) |
| 4 (former daily) | SMK_208 | Cigarettes per day (when smoked daily) |
| 3, 5, 6 (never daily) | — | NA::a (not applicable) |

### Current vs former smoker CPD

For users who need the underlying source variables:

| Smoking status | Variable to use |
|----------------|-----------------|
| Daily smoker | SMK_204 (or cigs_per_day) |
| Former daily smoker | SMK_208 (or cigs_per_day) |
| Occasional smoker | SMK_05B, SMK_05C |
| Former occasional | SMK_05B (at time smoked) |

## Era mappings

| Variable | Era 1 (2001-2014) | Era 2 (2015-2021) | Era 3 (2022+) |
|----------|-------------------|-------------------|---------------|
| SMK_204 | SMKA/C/E_204, SMK_204 | SMK_045 | CSS_25 |
| SMK_208 | SMKA/C/E_208, SMK_208 | SMK_075 | SPU_20 |
| SMK_05B | SMKA/C/E_05B, SMK_05B | SMK_050 | CSS_30 |
| SMK_05C | SMKA/C/E_05C, SMK_05C | SMK_055 | CSS_35 |

## Valid ranges

### CPD ranges by file type

| Variable | PUMF valid range | Master valid range | Notes |
|----------|:----------------:|:------------------:|-------|
| SMK_204 | 1-50 | 1-150 | PUMF capped; Master extreme values rare but valid |
| SMK_208 | 1-50 | 1-150 | Same as SMK_204 |
| SMK_05B | 1-50 | 1-50 | Occasional: fewer per day than daily |
| SMK_05C | 1-30 | 1-30 | Days per month (capped at 30) |

### Plausibility considerations

| Value | Assessment | Recommended action |
|-------|------------|-------------------|
| CPD = 0 | Invalid for daily smokers | Set to NA(b) or exclude |
| CPD = 1-50 | Normal range | Accept |
| CPD = 51-99 | Heavy but plausible | Accept (Master only) |
| CPD = 100+ | Extreme, likely error | Flag for review |
| Days/month = 30+ | By definition not occasional | Check smoker type |

### CPD equivalent for occasional smokers

The derived CPD equivalent uses:

```
CPD_equivalent = (SMK_05B × SMK_05C) / 30
```

Valid range for CPD_equivalent: 0.03-50

- Minimum: 1 cigarette × 1 day ÷ 30 = 0.03
- Maximum: 50 cigarettes × 30 days ÷ 30 = 50 (but this would make them daily)

## Validation status

| Level | Status | Notes |
|-------|--------|-------|
| L0 | ✓ | Schema compliant |
| L1 | ✓ | Database names validated |
| L2 | ✓ | Source references verified |
| L3 | ✓ | Category codes verified |
| L4 | ✓ | Range validation reviewed |
| L5 | ✓ | Integration tests (below) |

::: {.callout-note}
## Production status

SMK_204, SMK_208, SMK_05B, and SMK_05C are already in cchsflow v2.x production. The unified `cigs_per_day` variable is new in v3.0.0.
:::

## Integration testing

This section validates the primary recommended variable `cigs_per_day` using actual PUMF data. Testing includes:

1. **rec_with_table()**: Validation that the derived variable produces expected results
2. **Universe validation**: Only ever-daily smokers (SMKDSTY 1, 2, 4) should have values
3. **Clinical reasonableness**: Distribution checks against epidemiological expectations

**Note**: `cigs_per_day` is a derived variable routing SMK_204 and SMK_208 based on smoking status. Ground truth comparison is possible by verifying the routing logic.

### Clinical expectations

| Metric | Expected value |
|--------|----------------|
| Mean CPD | 12-20 cigarettes/day |
| Median CPD | 10-15 cigarettes/day |
| Range | 1-50 (PUMF capped) |
| Universe | Ever-daily smokers only (SMKDSTY 1, 2, 4) |

**Epidemiological context**: Typical daily smokers smoke 10-20 cigarettes per day (half to one pack). PUMF caps at 50 CPD.

```{r}
#| label: setup-integration-intensity
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)
library(here)

# Project paths
project_root <- here::here()
rdata_dir <- file.path(project_root, "working_data_and_documentation/pumf-rdata")

# Source cchsflow functions
original_wd <- getwd()
setwd(project_root)
source("R/strings.R")
source("R/recode-with-table.R")
setwd(original_wd)

# Load intensity worksheets
intensity_variables <- read.csv(file.path(project_root, "ceps/cep-002-smoking/04-intensity/variables_smk_intensity.csv"), stringsAsFactors = FALSE)
intensity_variable_details <- read.csv(file.path(project_root, "ceps/cep-002-smoking/04-intensity/variable_details_smk_intensity.csv"), stringsAsFactors = FALSE)

# Also need status variables for SMKDSTY_cat6
status_variables <- read.csv(file.path(project_root, "ceps/cep-002-smoking/01-status/variables_smk_status.csv"), stringsAsFactors = FALSE)
status_variable_details <- read.csv(file.path(project_root, "ceps/cep-002-smoking/01-status/variable_details_smk_status.csv"), stringsAsFactors = FALSE)

all_variables <- rbind(intensity_variables, status_variables)
all_variable_details <- rbind(intensity_variable_details, status_variable_details)

# Cycle order
cycle_order <- c(
  "2001", "2003", "2005", "2007_2008", "2009_2010",
  "2011_2012", "2013_2014", "2015_2016", "2017_2018",
  "2019_2020", "2022"
)
```

### rec_with_table() validation

Test that `cigs_per_day` produces continuous values in the expected range for ever-daily smokers.

```{r}
#| label: rec-with-table-intensity
#| code-fold: true

# Test rec_with_table() for cigs_per_day across cycles
rwt_results <- list()

for (cycle in cycle_order) {
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) next

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  db_name <- paste0("cchs", cycle, "_p")

  tryCatch({
    harmonized <- rec_with_table(
      data = df,
      variables = all_variables,
      variable_details = all_variable_details,
      database_name = db_name,
      var_names = c("cigs_per_day", "SMKDSTY_cat6"),
      log = FALSE
    )

    if ("cigs_per_day" %in% names(harmonized)) {
      cpd_values <- harmonized$cigs_per_day
      status_values <- harmonized$SMKDSTY_cat6

      # Filter to ever-daily smokers (SMKDSTY 1, 2, 4)
      ever_daily <- status_values %in% c(1, 2, 4)
      cpd_ever_daily <- cpd_values[ever_daily]
      cpd_valid <- cpd_ever_daily[!is.na(cpd_ever_daily)]

      # Calculate statistics
      if (length(cpd_valid) > 0) {
        rwt_results[[cycle]] <- data.frame(
          Cycle = cycle,
          N = nrow(df),
          N_ever_daily = sum(ever_daily, na.rm = TRUE),
          N_valid = length(cpd_valid),
          Mean = round(mean(cpd_valid), 1),
          Median = round(median(cpd_valid), 1),
          Min = round(min(cpd_valid), 0),
          Max = round(max(cpd_valid), 0),
          Pct_valid = round(100 * length(cpd_valid) / sum(ever_daily, na.rm = TRUE), 1),
          check.names = FALSE
        )
      }
    }
  }, error = function(e) {
    message("rec_with_table failed for ", cycle, ": ", e$message)
  })
}

rwt_df <- do.call(rbind, rwt_results)
rownames(rwt_df) <- NULL
```

```{r}
#| label: tbl-intensity-results
#| tbl-cap: "rec_with_table() results: cigs_per_day across PUMF cycles"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  kable(rwt_df,
        col.names = c("Cycle", "N Total", "Ever Daily", "Valid", "Mean", "Median", "Min", "Max", "% Valid"),
        format.args = list(big.mark = ","))
}
```

### Universe validation

Verify that `cigs_per_day` is NA for never-daily smokers (SMKDSTY 3, 5, 6).

```{r}
#| label: universe-validation-intensity
#| code-fold: true

# Check that never-daily smokers have NA
universe_check <- list()

for (cycle in c("2007_2008", "2015_2016")) {  # Sample cycles from each era
  rdata_file <- file.path(rdata_dir, paste0("CCHS_", cycle, ".RData"))

  if (!file.exists(rdata_file)) next

  env <- new.env()
  load(rdata_file, envir = env)
  df <- get(ls(env)[1], envir = env)

  db_name <- paste0("cchs", cycle, "_p")

  tryCatch({
    harmonized <- rec_with_table(
      data = df,
      variables = all_variables,
      variable_details = all_variable_details,
      database_name = db_name,
      var_names = c("cigs_per_day", "SMKDSTY_cat6"),
      log = FALSE
    )

    status <- harmonized$SMKDSTY_cat6
    cpd <- harmonized$cigs_per_day

    # Count by status
    for (s in 1:6) {
      in_status <- status == s & !is.na(status)
      n_in_status <- sum(in_status, na.rm = TRUE)
      n_has_cpd <- sum(!is.na(cpd[in_status]), na.rm = TRUE)
      pct_has_cpd <- if (n_in_status > 0) round(100 * n_has_cpd / n_in_status, 1) else NA

      universe_check[[paste0(cycle, "_", s)]] <- data.frame(
        Cycle = cycle,
        Status = s,
        Status_label = c("Daily", "Occ (fmr daily)", "Always occ", "Fmr daily", "Fmr occ", "Never")[s],
        N_status = n_in_status,
        N_has_cpd = n_has_cpd,
        Pct_has_cpd = pct_has_cpd,
        Expected = ifelse(s %in% c(1, 2, 4), "Has CPD", "NA"),
        check.names = FALSE
      )
    }
  }, error = function(e) {
    message("Universe check failed for ", cycle, ": ", e$message)
  })
}

universe_df <- do.call(rbind, universe_check)
rownames(universe_df) <- NULL
```

```{r}
#| label: tbl-universe-intensity
#| tbl-cap: "Universe validation: cigs_per_day by smoking status"

if (!is.null(universe_df) && nrow(universe_df) > 0) {
  kable(universe_df,
        col.names = c("Cycle", "Status", "Label", "N", "Has CPD", "% Has CPD", "Expected"))
}
```

### Clinical assessment

```{r}
#| label: clinical-assessment-intensity
#| code-fold: true

assessment <- list()

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  # Check 1: Mean CPD in expected range
  mean_range <- range(rwt_df$Mean, na.rm = TRUE)
  check_mean <- mean_range[1] >= 10 && mean_range[2] <= 25
  assessment$mean_range <- list(
    check = "Mean CPD 10-25",
    observed = paste0(mean_range[1], " - ", mean_range[2], " cigs/day"),
    pass = check_mean
  )

  # Check 2: Median around 10-20
  median_range <- range(rwt_df$Median, na.rm = TRUE)
  check_median <- median_range[1] >= 8 && median_range[2] <= 25
  assessment$median_range <- list(
    check = "Median CPD 8-25",
    observed = paste0(median_range[1], " - ", median_range[2], " cigs/day"),
    pass = check_median
  )

  # Check 3: Valid percentage high
  pct_valid_range <- range(rwt_df$Pct_valid, na.rm = TRUE)
  check_coverage <- pct_valid_range[1] >= 85
  assessment$coverage <- list(
    check = "Coverage ≥85% of ever-daily smokers",
    observed = paste0(pct_valid_range[1], "% - ", pct_valid_range[2], "%"),
    pass = check_coverage
  )

  # Check 4: Universe correctness (never-daily should have low %)
  if (exists("universe_df") && nrow(universe_df) > 0) {
    never_daily <- universe_df[universe_df$Status %in% c(3, 5, 6), ]
    max_pct_never_daily <- max(never_daily$Pct_has_cpd, na.rm = TRUE)
    check_universe <- max_pct_never_daily < 5
    assessment$universe <- list(
      check = "Never-daily smokers have <5% with CPD",
      observed = paste0("Max: ", max_pct_never_daily, "%"),
      pass = check_universe
    )
  }

  # Check 5: Values within PUMF bounds (1-50)
  max_value <- max(rwt_df$Max, na.rm = TRUE)
  min_value <- min(rwt_df$Min, na.rm = TRUE)
  check_bounds <- min_value >= 1 && max_value <= 50
  assessment$bounds <- list(
    check = "Values within [1, 50] (PUMF bounds)",
    observed = paste0(min_value, " - ", max_value, " cigs/day"),
    pass = check_bounds
  )
}

assessment_df <- do.call(rbind, lapply(assessment, function(x) {
  data.frame(Check = x$check, Observed = x$observed,
             Result = ifelse(x$pass, "✓ PASS", "✗ FAIL"))
}))
```

```{r}
#| label: tbl-assessment-intensity
#| tbl-cap: "Clinical reasonableness assessment"

if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) {
  kable(assessment_df, row.names = FALSE)
}
```

### CPD distribution trend

```{r}
#| label: fig-cpd-trend
#| fig-cap: "Mean cigarettes per day by cycle"

if (!is.null(rwt_df) && nrow(rwt_df) > 0) {
  years <- as.numeric(gsub("_.*", "", rwt_df$Cycle))

  plot(years, rwt_df$Mean, type = "b", pch = 19, col = "steelblue",
       xlab = "CCHS Year", ylab = "Mean Cigarettes per Day",
       main = "Smoking Intensity Trend",
       ylim = c(10, 25))

  lines(years, rwt_df$Median, type = "b", pch = 17, col = "darkgreen", lty = 2)

  legend("topright", legend = c("Mean", "Median"),
         col = c("steelblue", "darkgreen"), pch = c(19, 17), lty = c(1, 2))
}
```

### Test summary

```{r}
#| label: test-summary-intensity

n_clinical <- if (exists("assessment_df") && !is.null(assessment_df)) nrow(assessment_df) else 0
n_clinical_pass <- if (exists("assessment_df") && !is.null(assessment_df) && nrow(assessment_df) > 0) sum(grepl("PASS", assessment_df$Result)) else 0
n_cycles <- if (!is.null(rwt_df)) nrow(rwt_df) else 0

cat("## cigs_per_day Integration Test Summary\n\n")
cat("**Clinical reasonableness:**", n_clinical_pass, "/", n_clinical, "passed\n")
cat("**Cycles tested:**", n_cycles, "\n")

if (n_cycles == 0) {
  cat("\n⚠ **No PUMF data available for testing**\n")
} else if (n_clinical == 0 || n_clinical_pass == n_clinical) {
  cat("\n✓ **L5 INTEGRATION TESTS: PASSED**\n")
} else {
  cat("\n✗ **L5 INTEGRATION TESTS: ISSUES DETECTED**\n")
}
```

## Appendix

- [L0: Documentation review](04-intensity/L0_documentation_review.md)
- [L1: Concordance](04-intensity/L1_concordance.md)
- [L2: Gap analysis](04-intensity/L2_gap_analysis.md)
- [L3: Worksheets](04-intensity/L3_worksheets.md)
- [L4: Derived variables](04-intensity/L4_derived_variables.md)
- [variables_smk_intensity.csv](04-intensity/variables_smk_intensity.csv)
- [variable_details_smk_intensity.csv](04-intensity/variable_details_smk_intensity.csv)
