---
title: "05 - Pack-Years Variables"
author: "Doug Manuel, Maikol Diasparra, Caitlin St-Onge"
date: "2026-01-08"
format:
  html:
    toc: true
    toc-depth: 2
---

## Overview

| Attribute | Value |
|-----------|-------|
| **Subgroup** | 05-pack-years |
| **Variables** | 2 |
| **Purpose** | Cumulative smoking exposure for dose-response analysis |
| **Dependencies** | 01-status, 02-initiation, 03-cessation, 04-intensity |
| **Dependents** | None (end-point variable) |

Pack-years is the key derived variable for smoking research. It combines information from all other subgroups to calculate cumulative exposure.

## Variables

```{r}
#| label: tbl-packyears-variables
#| echo: false
#| message: false
#| warning: false

source("../../R/table-generators.R")

generate_variable_list_table("05-pack-years/variables_smk_packyears.csv") |>
  knitr::kable(align = c("l", "l", "l", "c", "c", "l"))
```

## PUMF cycle coverage

```{r}
#| label: tbl-packyears-coverage
#| echo: false
#| message: false
#| warning: false

generate_subject_coverage_matrix("05-pack-years") |>
  knitr::kable(align = c("l", rep("c", 13)))
```

## Formula

```
pack_years = (cigarettes_per_day / 20) × years_smoked
```

Where:
- `cigarettes_per_day` = `cigs_per_day` (unified variable, routes SMK_204/SMK_208 by status)
- `years_smoked` = current_age - age_started - time_quit (if applicable)

## Input variables

| Component | Unified variable | Notes |
|-----------|------------------|-------|
| Smoking status | SMKDSTY_A | 6-category status determines routing |
| Age started | age_start_smoking | Routes SMK_040/SMKG040_cont by file type |
| CPD (daily) | cigs_per_day | Routes SMK_204/SMK_208 by status |
| Time since quit | **time_quit_complete** | All former smokers (correctly handles SMK_10_gate) |
| Current age | DHHGAGE_cont | PUMF grouped → continuous via midpoint |

::: {.callout-tip}
## Unified input variables

The unified variables (`age_start_smoking`, `cigs_per_day`, `time_quit_complete`) automatically route to the correct source variable based on SMKDSTY_A smoking status and file type. This simplifies the pack-years calculation interface.
:::

::: {.callout-warning}
## Use `time_quit_complete`, not `time_quit_smoking`

**`time_quit_complete`** correctly handles the SMK_10_gate for "reducers" (former daily smokers who continued occasional before quitting completely). The older `time_quit_smoking` variable gives incorrect values for reducers - it uses SMK_09A (when stopped daily) instead of SMK_10A (when quit completely).

**Availability:** `time_quit_complete` requires SMK_10_gate, which was not collected in 2001. For 2001-only analyses, use `time_quit_smoking` with documented limitation.
:::

## PUMF vs Master precision

::: {.callout-warning}
## Precision difference

PUMF pack-years estimates have ~15-20% relative error compared to Master due to:

| Input | PUMF error | Master error |
|-------|------------|--------------|
| Age | ~±2.5 years (grouped) | None |
| Age started | ~±3 years (midpoint) | None |
| CPD | Capped at 50 | None |
| Time quit | ~±1.5 years (midpoint) | ~±0.5 years |
:::

### Error propagation example

For a 50-year-old who started at 18, smokes 25 CPD, and quit at 45:

| Source | Years smoked | Pack-years |
|--------|--------------|------------|
| Master (true) | 27 years | 33.75 |
| PUMF (estimate) | 24-30 years | 30-38 |
| Relative error | ~±11% | ~±12% |

## Derivation logic

```r
pack_years_der <- function(data, ...) {

 # Step 1: Determine smoking status
 # - Current daily: calculate ongoing exposure
 # - Current occasional: calculate with adjusted CPD
 # - Former daily: calculate past exposure
 # - Former occasional: calculate past exposure
 # - Never: return 0

 # Step 2: Calculate years smoked
 # - Current: age - age_started
 # - Former: age - age_started - time_quit

 # Step 3: Calculate pack-years
 # - (cpd / 20) × years_smoked

 # Step 4: Validate output
 # - Must be >= 0
 # - Flag if > 165 (theoretical max: 3 packs/day × 55 years)
}
```

## Special cases

### Never smokers

| Output | Value |
|--------|-------|
| pack_years_der | 0 |
| smoking_years_der | 0 |

### Occasional smokers

```
CPD_equivalent = (SMK_05B × SMK_05C) / 30
pack_years = (CPD_equivalent / 20) × years_smoked
```

### Former occasional (never daily)

Limited data available. Uses:
- SMKG06C for quit year (when available)
- Conservative estimates for duration

### Missing inputs

If any required input is missing:
- Return NA with provenance tag indicating which input was missing
- Do not impute missing values

## Valid ranges and validation

### Output ranges

| Variable | Min | Max | Typical | Derivation of max |
|----------|:---:|:---:|:-------:|-------------------|
| pack_years_der | 0 | 165 | 0-60 | 3 packs/day × 55 years |
| smoking_years_der | 0 | 80 | 0-50 | Age minus minimum start age |

### Input ranges (enforced)

| Input | Valid range | Source |
|-------|-------------|--------|
| age_started | 5-80 | 02-initiation |
| cpd | 1-150 | 04-intensity |
| time_quit | 0-80 | 03-cessation |
| current_age | 12-100 | CCHS respondent age |

### Cross-field constraints

| Constraint | Formula | Violation action |
|------------|---------|------------------|
| Years smoked ≥ 0 | age - age_started - time_quit ≥ 0 | Set to NA(b) |
| Years smoked ≤ age - 5 | years_smoked ≤ age - 5 | Cap at maximum |
| Pack-years plausible | pack_years ≤ years_smoked × 7.5 | Flag for review |
| CPD consistent | CPD > 0 for ever-smokers | Set to NA(b) |

The pack-years plausibility check assumes maximum 3 packs/day (150 CPD ÷ 20 = 7.5 packs).

### Quality flags

Pack-years calculation sets quality flags for:

| Condition | Severity | Action |
|-----------|----------|--------|
| Age started > current age | Error | Set to NA(b) |
| Time quit > years smoked | Error | Set to NA(b) |
| CPD > 60 | Warning | Keep value, flag |
| Pack-years > 100 | Warning | Keep value, flag |
| Missing required input | Error | Set to NA with provenance |

### Distribution expectations

Based on CCHS data, expected pack-years distribution:

| Pack-years range | Expected % | Notes |
|------------------|:----------:|-------|
| 0 | ~50-60% | Never smokers |
| 0.1-10 | ~15-20% | Light/short-term |
| 10-30 | ~10-15% | Moderate |
| 30-60 | ~5-10% | Heavy |
| 60+ | ~1-3% | Very heavy |

Values significantly outside this distribution may indicate data quality issues.

## Era-specific considerations

### 2001-2014 (Era 1)

Full PUMF support using SMKG203_cont/SMKG207_cont for age started.

### 2015-2021 (Era 2)

Age started variables dropped from PUMF. Options:
1. Use Master files (SMK_040)
2. Use SMKG01C_cont (age first cigarette) as proxy
3. Accept reduced precision

### 2022+ (Era 3)

Module split. Master files have CSS_*/SPU_* equivalents.

## Validation status

| Level | Status | Notes |
|-------|--------|-------|
| L0 | ✓ | Schema compliant |
| L1 | ✓ | Database names validated |
| L2 | ✓ | Source references verified |
| L3 | ✓ | Formula verified |
| L4 | ✓ | Edge cases documented |
| L5 | ✓ | Integration tests (below) |

## Integration testing

::: {.callout-note}
## Integration testing requires full cchsflow package

The `pack_years_der` variable is a complex derived variable that requires the full cchsflow package infrastructure. Integration tests for this variable are available in the [appendix-integration-test.qmd](appendix-integration-test.qmd) which uses the complete package setup.

**Expected validation criteria:**

| Check | Expected result |
|-------|-----------------|
| Mean pack-years | 3-20 (population-wide, includes never smokers) |
| Median pack-years | 0 (majority are never smokers) |
| Max pack-years | ≤165 (theoretical bound) |
| % with zero | 40-70% (never smokers) |
| % with >60 PY | ≤5% (very heavy smokers) |

**By smoking status expectations:**

- Never smokers (SMKDSTY 6): pack_years = 0
- Former smokers (SMKDSTY 4, 5): pack_years > 0
- Current smokers (SMKDSTY 1, 2, 3): pack_years > 0 (ongoing exposure)
:::

## Appendix

- [L0: Documentation assessment](05-pack-years/L0_documentation_assessment.md)
- [L1: Variable concordance](05-pack-years/L1_variable_concordance.md)
- [L2: Semantic mapping](05-pack-years/L2_semantic_mapping.md)
- [L3: Worksheet authoring](05-pack-years/L3_worksheet_authoring.md)
- [L4: DV specifications](05-pack-years/L4_dv_specifications.md)
- [variables_smk_packyears.csv](05-pack-years/variables_smk_packyears.csv)
- [variable_details_smk_packyears.csv](05-pack-years/variable_details_smk_packyears.csv)
