---
title: "CEP-007: Diet variable cross-cycle prevalence"
author: "cchsflow review (PR #148)"
date: "2026-02-11"
format:
  html:
    toc: true
    code-fold: true
---

## Overview

PR #148 adds master (`_m`) database support and 2017-2018 coverage for 9 diet
variables: FVCDFRU, FVCDSAL, FVCDPOT, FVCDCAR, FVCDVEG, FVCDJUI (continuous
inputs), diet_score (derived), and diet_score_cat3 (categorical derived).

This document visualises the L6 integration test results from
`rec_with_table()` run against all PUMF cycles. The purpose is to detect
era boundary harmonization errors — sudden shifts in exposure distributions
at the 2007 or 2015 transitions that would signal a mapping or naming problem.

**Known data patterns:**

- **2005**: FVC was optional content (BC, ON, AB, PE only) — expect ~56% valid
- **2017-2018**: FVC was optional content — expect ~1% valid
- **2015 variable rename**: FVCDFRU → FVCDVFRU, FVCDSAL → FVCDVGRN, etc.

```{r}
#| label: setup
#| message: false
#| warning: false

results <- read.csv("diet-pumf-integration-test.csv", stringsAsFactors = FALSE)

# Extract start year from cycle name for x-axis
results$year <- as.numeric(gsub("cchs(\\d{4}).*", "\\1", results$cycle))

# Keep only the main biennial cycles (drop annual: 2010, 2012, 2014)
main_cycles <- c("cchs2001_p", "cchs2003_p", "cchs2005_p",
                 "cchs2007_2008_p", "cchs2009_2010_p",
                 "cchs2011_2012_p", "cchs2013_2014_p",
                 "cchs2015_2016_p", "cchs2017_2018_p")
main <- results[results$cycle %in% main_cycles, ]
```

## FVC input variables: valid % across cycles

Each FVC input variable should track closely. A divergence between variables
within the same cycle would indicate a variable-specific mapping error.

```{r}
#| label: fig-fvc-valid-pct
#| fig-cap: "Valid % for FVC input variables across CCHS PUMF cycles"
#| fig-width: 9
#| fig-height: 5

fvc_vars <- c("FVCDFRU", "FVCDSAL", "FVCDPOT", "FVCDCAR", "FVCDVEG", "FVCDJUI")
fvc_colours <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00", "#a65628")

plot(NULL, xlim = c(2001, 2018), ylim = c(0, 105),
     xlab = "CCHS cycle start year", ylab = "Valid %",
     main = "FVC input variables: cross-cycle valid %",
     xaxt = "n")
axis(1, at = unique(main$year), labels = unique(main$year), las = 2)

# Era boundary reference lines
abline(v = 2007, lty = 2, col = "grey60")
abline(v = 2015, lty = 2, col = "grey60")
text(2007, 103, "pre-2007 era", pos = 4, cex = 0.7, col = "grey40")
text(2015, 103, "2015+ era", pos = 4, cex = 0.7, col = "grey40")

for (i in seq_along(fvc_vars)) {
  v <- fvc_vars[i]
  d <- main[main$variable == v, ]
  d <- d[order(d$year), ]
  lines(d$year, d$valid_pct, type = "b", pch = 19, col = fvc_colours[i], lwd = 1.5)
}

legend("bottomleft", legend = fvc_vars, col = fvc_colours,
       lwd = 1.5, pch = 19, cex = 0.7, bg = "white")
```

**Interpretation:** All six FVC variables track together across cycles, which is
expected since they come from the same survey module. The 2005 dip (~57-60%) and
2017-2018 dip (~1%) reflect optional content — not harmonization errors. The
2014 → 2015 transition is clean with no step change, confirming the 2015+
variable renames (FVCDVFRU, FVCDVGRN, etc.) are correctly mapped.

## diet_score: cross-cycle valid %

The derived `diet_score` should have lower valid % than individual FVC inputs
(since all inputs must be non-missing for the score to compute), but should
follow the same cross-cycle pattern.

```{r}
#| label: fig-diet-score-valid
#| fig-cap: "diet_score valid % compared to FVCDFRU input"
#| fig-width: 9
#| fig-height: 5

ds <- main[main$variable == "diet_score", ]
ds <- ds[order(ds$year), ]
fru <- main[main$variable == "FVCDFRU", ]
fru <- fru[order(fru$year), ]

plot(ds$year, ds$valid_pct, type = "b", pch = 19, col = "#e41a1c", lwd = 2,
     xlim = c(2001, 2018), ylim = c(0, 105),
     xlab = "CCHS cycle start year", ylab = "Valid %",
     main = "diet_score vs FVCDFRU: cross-cycle valid %",
     xaxt = "n")
axis(1, at = unique(ds$year), labels = unique(ds$year), las = 2)
lines(fru$year, fru$valid_pct, type = "b", pch = 17, col = "#377eb8", lwd = 1.5)

abline(v = c(2007, 2015), lty = 2, col = "grey60")

legend("bottomleft", legend = c("diet_score", "FVCDFRU (input)"),
       col = c("#e41a1c", "#377eb8"), pch = c(19, 17), lwd = c(2, 1.5),
       cex = 0.8, bg = "white")
```

**Interpretation:** diet_score tracks slightly below its FVC inputs (as expected
— all 6 inputs must be valid). The gap is small (1-6 percentage points),
confirming the DV function is not excessively dropping valid cases.

## diet_score_cat3: category distribution across cycles

The most important harmonization check for the categorical derived variable.
A sudden shift in the poor/fair/adequate distribution at an era boundary would
signal a recoding or mapping error.

```{r}
#| label: fig-diet-cat-distribution
#| fig-cap: "diet_score_cat3 distribution across cycles (excluding NA)"
#| fig-width: 9
#| fig-height: 6

# Load the full integration test data to get category distributions
# We need to rerun rec_with_table to get actual categories, since the CSV
# only has valid counts. Use the stored diet_score valid % to label cycles.

# For this visualisation, we use the diet_score valid % and the known
# category distribution from the PR-148 review summary:
# Poor (1): 0-2, Fair (2): 2-8, Adequate (3): 8-10

cat_data <- data.frame(
  cycle = c("2001", "2003", "2005", "2007", "2009", "2011", "2013", "2015", "2017"),
  poor = c(43, 24, 11, 26, 21, 15, 14, 7, 0),
  fair = c(145, 139, 92, 137, 142, 128, 136, 161, 2),
  adequate = c(11, 25, 9, 26, 24, 34, 32, 23, 0),
  stringsAsFactors = FALSE
)

# Calculate proportions (among valid respondents only)
cat_data$total <- cat_data$poor + cat_data$fair + cat_data$adequate
cat_data$pct_poor <- round(100 * cat_data$poor / cat_data$total, 1)
cat_data$pct_fair <- round(100 * cat_data$fair / cat_data$total, 1)
cat_data$pct_adequate <- round(100 * cat_data$adequate / cat_data$total, 1)

# Exclude 2017-2018 from the bar chart (only 2 valid respondents)
cat_plot <- cat_data[cat_data$cycle != "2017", ]

# Stacked bar chart
bar_matrix <- rbind(cat_plot$pct_poor, cat_plot$pct_fair, cat_plot$pct_adequate)
colnames(bar_matrix) <- cat_plot$cycle

barplot(bar_matrix,
        col = c("#d73027", "#fee08b", "#1a9850"),
        border = NA,
        main = "diet_score_cat3: exposure distribution across cycles\n(% of valid respondents, N=200 per cycle)",
        ylab = "Percentage",
        xlab = "CCHS cycle start year",
        ylim = c(0, 110),
        las = 1)

# Add era boundary markers
abline(v = 3.7, lty = 2, col = "grey40")   # Between 2005 and 2007
abline(v = 8.5, lty = 2, col = "grey40")   # Between 2013 and 2015

legend("topright",
       legend = c("Adequate (8-10)", "Fair (2-8)", "Poor (0-2)"),
       fill = c("#1a9850", "#fee08b", "#d73027"),
       border = NA, cex = 0.8, bg = "white")
```

**Interpretation:** The distribution is stable across cycles — fair diet (score
2-8) dominates at ~70-80%, consistent with the population distribution. There is
no step change at the 2015 era boundary. The gradual increase in "adequate"
from 2001 (5.5%) to 2013 (17.6%) may reflect real dietary trends.
The 2005 cycle is excluded from trend interpretation due to optional content
(only 112 valid respondents from 4 provinces). The 2017-2018 cycle is excluded
entirely (only 2 valid respondents).

## Summary

| Check | Result |
|-------|--------|
| Era boundary (2015) | **Clean** — no step change in valid % or distributions |
| Era boundary (2007) | **Clean** — pre-2007 cycle letter mappings correct |
| FVC input consistency | **PASS** — all 6 inputs track together |
| DV completeness gap | **Acceptable** — diet_score 1-6% below inputs |
| Category stability | **Stable** — gradual trends, no abrupt shifts |
| Optional content | **Expected** — 2005 and 2017-2018 dips documented |

**Issue found:** 30 raw FVC variables (FVC_1A through FVC_6E) have `chs2011_2012_m`
and `chs2013_2014_m` typos (missing leading `c`). See
[PR-148-review-summary.md](PR-148-review-summary.md) for details.
